---
title: "Take-home Exercise 2: Anatomy of a Broken Embargo"
subtitle: "A visual-forensic reconstruction of the TenantThread leak (VAST Challenge 2026, Mini-Challenge 1)"
author: "Meng Tingyu"
date: today
date-modified: "last-modified"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show the code"
    code-tools: true
    code-overflow: wrap
    fig-width: 9
    fig-height: 6
    fig-align: center
execute:
  warning: false
  message: false
  freeze: auto
editor: visual
---

# Setting the scene {#sec-scene}

I am writing this as if I had been pulled in to help CivicLoom Realty Partners' legal team. CivicLoom had quietly signed a merger with TenantThread under a strict embargo that was supposed to hold until 6:00 PM on 5 June 2046. At around 5:00 PM, embargoed details of the deal started appearing on the social platform FleX, posted by accounts tied to TenantThread's own automated communications system. The embargo broke one hour before it was due to lift.

The legal team's question is blunt: **did TenantThread's agents deliberately leak the merger, or did the compliance system simply break down under pressure?** That framing invites a yes or no answer, and I think the honest finding sits in between. So my goal for this exercise is not to pick a side too quickly. It is to let the data narrate the two weeks leading up to the breach, and then to argue, from what I can actually see, what kind of failure this was.

My working thesis, which I will defend with evidence by the end, is this. The embargo was not hacked and it did not crash. It was talked past. Under genuine existential pressure, a collapsing stock and a deal about to die, the agents reframed a leak into a sequence of locally defensible actions, and the automated monitor was argued into co-signing each step. The embargo died of a thousand reasonable cuts.

::: callout-note
## A note on the data I received

The challenge package contained one substantive data file, `MC1_final_00.json`. The other items in the bundle turned out to be empty macOS resource stubs rather than the official data dictionary, so I reconstructed the schema directly from the JSON. Everything below is derived only from that file.
:::

# Getting to know the data {#sec-data}

`MC1_final_00.json` is a single object with one key, `rounds`. There are 23 rounds. Each round is one hour of simulated office life. The first 13 rounds are spaced one business day apart and cover the run-up from 17 May to 4 June. The remaining 10 rounds are hourly and cover the crisis itself on 5 June. Each round holds three blocks.

The `environment_context` block is the world state for that hour, a prose `event_narrative`, a short `event_headline`, a `market_snapshot` with the stock price and market sentiment, and lists such as `agents_unavailable` and `critical_deadlines`.

The `communications` block is the heart of the file, with 912 messages in total. Every message carries a sender, a `channel`, a `message_type`, a `responding_to` pointer that lets me thread conversations into a reply graph, the `content`, and, for a subset of messages, an `internal_state` object. That last field is the one I suspect most analysts will walk past, and it is the most valuable thing in the dataset. It records what the agent was privately thinking, split into `reacting`, `rationalizing`, and `deliberating`. In other words, the file does not just tell me what was said in public. It tells me how the decision was justified in private.

The `participants` block lists who was present each round and tags each one with a `declared_action` and a sentiment.

Seven agents appear: a `Legal-Agent`, a `Platform-Trust-Agent`, a `PR-Agent`, a `Social-Manager-Agent`, two interns, and a `Judge-Agent`. The Judge is the automated compliance monitor referred to in the brief as "The Judge". Six channels appear, and the distinction between them matters a great deal later: `comms_huddle` is the monitored team room, `one_on_one_chat` is private messaging, `side_huddle` is the back channel that the brief calls the Shadow channel, and `official_post`, `personal_post`, and `anonymous_post` are the three ways a message can reach the public.

# Data preparation and wrangling {#sec-wrangle}

I read the JSON without auto simplification so I could control exactly how the nested lists became tidy tables. Then I flattened it into three tibbles, one row per message, one row per round of world state, and a small lookup for the reply graph. The pattern below is the workhorse of the whole document.

```{r}
#| label: setup-load
pacman::p_load(tidyverse, jsonlite, ggraph, tidygraph,
               ggiraph, ggrepel, scales, knitr, tidytext)

# safe null-coalescing operator (works on every R version)
`%||%` <- function(a, b) if (is.null(a)) b else a

raw    <- fromJSON("data/MC1_final_00.json", simplifyVector = FALSE)
rounds <- raw$rounds

# One tidy row per message --------------------------------------------------
messages <- imap(rounds, function(rd, i) {
  map(rd$communications, function(m) {
    st <- m$internal_state %||% list()
    tibble(
      round         = i - 1L,
      hour          = rd$hour,
      message_id    = m$message_id    %||% NA_character_,
      agent_label   = m$agent_label   %||% NA_character_,
      agent_role    = m$agent_role    %||% NA_character_,
      channel       = m$channel       %||% NA_character_,
      message_type  = m$message_type  %||% NA_character_,
      responding_to = m$responding_to %||% NA_character_,
      reacting      = st$reacting      %||% NA_character_,
      rationalizing = st$rationalizing %||% NA_character_,
      deliberating  = st$deliberating  %||% NA_character_,
      content       = m$content       %||% NA_character_
    )
  }) |> list_rbind()
}) |> list_rbind()

dim(messages)
```

The world-state table needed more care, because the `stock_price` field is dirty in a very instructive way. The pre-crisis values are clean dollar strings that slide gently downward. The crisis-day values are a mess: two rounds report the literal string `NaN`, one round reports `$180` (clearly not a real intraday price for a stock that opened around thirty dollars), and two afternoon rounds report `$18.0`, which is suspicious because that is exactly the figure the narrative gives for annual recurring revenue of eighteen million. My read is that the price field was contaminated with the ARR number during the panic. So I parse what I can, set the rest to missing, and I do not pretend the broken crisis prices are a smooth series.

```{r}
#| label: wrangle-env
clean_num <- function(x, strip = "\\$") {
  x |>
    str_remove_all(strip) |>
    str_remove_all("%") |>
    str_trim() |>
    na_if("NaN") |> na_if("None") |> na_if("") |>
    as.numeric()
}

environment <- imap(rounds, function(rd, i) {
  ec <- rd$environment_context %||% list()
  ms <- ec$market_snapshot     %||% list()
  un <- unlist(ec$agents_unavailable %||% list())
  tibble(
    round         = i - 1L,
    hour          = rd$hour,
    headline      = ec$event_headline   %||% NA_character_,
    narrative     = ec$event_narrative  %||% NA_character_,
    stock_raw     = ms$stock_price       %||% NA_character_,
    sentiment     = ms$sentiment         %||% NA_character_,
    pct_change    = clean_num(ms$percent_change %||% NA_character_),
    n_unavailable = length(un),
    judge_offline = "Judge" %in% un
  )
}) |> list_rbind()

# Parse the price, then flag the values I do not trust ----------------------
environment <- environment |>
  mutate(
    stock_clean = clean_num(stock_raw),
    # treat the $180 glitch and the $18.0 ARR contamination as missing
    stock_valid = if_else(round %in% c(15, 18, 19) | stock_clean > 50,
                          NA_real_, stock_clean)
  )
```

Because the price collapses into noise on the worst day, I needed a signal that is complete across all 23 rounds to act as the backbone of my timeline. Market sentiment fits. It is recorded every round, so I encoded it onto a simple five-point stress scale. This keeps the pressure story honest: I am charting the mood of the market, which is fully observed, rather than a price that is half missing.

```{r}
#| label: wrangle-stress
stress_map <- c(neutral = 1, cautious = 2, negative = 3,
                critical = 4, LOW = 4, CRITICAL = 5, RECOVERING = 3)

environment <- environment |>
  mutate(stress = unname(stress_map[sentiment]),
         phase  = if_else(round <= 12, "Pre-crisis (daily)", "Crisis day (hourly)"))

# axis labels reused by several charts below: dates for pre-crisis, times for crisis day
x_lab <- environment |>
  mutate(lab = if_else(round <= 12, str_sub(hour, 6, 10), str_sub(hour, 12, 16)))

environment |>
  select(round, hour, stock_raw, stock_valid, sentiment, stress, n_unavailable) |>
  kable(caption = "World-state table after cleaning. Note how the price field degrades into NaN and contaminated values on 5 June, while sentiment stays usable throughout.")
```

::: callout-warning
## The most important thing the cleaning revealed

The price data does not just go missing on crisis day. It gets *worse than missing*: the field starts carrying the wrong number. A reader who trusted `stock_price` blindly would have plotted an eighteen-dollar afternoon as if the company were already nearly worthless, when eighteen was actually the revenue figure that triggered a loan covenant breach. Cleaning was not cosmetic here. It changed the story.
:::

Once the price is clean, the pre-crisis half of the story is easy to see. I plotted the daily closing price for the two weeks before the crisis, with each day's percentage move printed underneath. This is the slow bleed that set the stage.

```{r}
#| label: fig-stockbleed
#| fig-cap: "The share price before crisis day. A steady two-week decline from $38.70 to $31.50, accelerating after the NHPI report and the first SaltWind piece. This is the pressure that was already building before anyone touched the embargo."
environment |>
  filter(round <= 12) |>
  ggplot(aes(round, stock_valid)) +
  annotate("rect", xmin = 2.5, xmax = 12.5, ymin = 30.5, ymax = 39.5,
           fill = "#fff4f2") +
  geom_line(colour = "#cb181d", linewidth = 1) +
  geom_point(colour = "#cb181d", size = 2.6) +
  geom_text(aes(label = scales::dollar(stock_valid)), vjust = -1.1,
            size = 2.9, colour = "grey25") +
  geom_text(aes(label = paste0(pct_change, "%"), y = 30.7),
            size = 2.6, colour = "#737373") +
  annotate("text", x = 3, y = 38.8, label = "NHPI report", size = 3,
           colour = "#cb181d", hjust = 0) +
  annotate("text", x = 10, y = 34.2, label = "SaltWind #1", size = 3,
           colour = "#cb181d", hjust = 1) +
  scale_x_continuous(breaks = 0:12,
                     labels = str_sub(environment$hour[1:13], 6, 10)) +
  scale_y_continuous(labels = scales::dollar_format(), limits = c(30.5, 39.5)) +
  labs(title = "A quiet, steady bleed for two weeks",
       subtitle = "Daily close (red) with each day's move (grey)",
       x = NULL, y = "Share price") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        panel.grid.minor = element_blank())
```

The decline is gentle but relentless, and it steepens exactly when the external pressure arrives. Nothing here is a crisis on its own. It is the accumulating strain that makes the crisis-day decisions feel, to the agents living through them, like emergencies that justify exceptions.

# Task 1: reconstructing the sequence {#sec-task1}

## The pressure timeline

My first job is to lay out the sequence of events and mark the decisions and system elements that let the post slip past the embargo. I built an interactive timeline. The grey backbone is market stress over the two weeks. The dots are the events that matter, colour coded by what kind of moment each one is. The pale band on the right is the crisis day. Hover any dot to read the full narrative for that hour.

I deliberately separated four kinds of moment, because the argument of this whole exercise is that the breach was not one event but a chain. There were quiet *deviations* from normal practice, one genuine *near-miss* leak, the *control* that was installed in response, the points where that *guardrail was bent*, the moment the *control failed* outright, and finally the *breach* itself.

```{r}
#| label: fig-timeline
#| fig-cap: "The road to the breach. Market stress (grey step line) climbs for two weeks while a chain of decisions quietly lowers the company's resistance to leaking. Hover the points for detail."
#| fig-height: 6.5
events <- tribble(
  ~round, ~label,                                   ~kind,
  3,  "Shadow channel first used",                  "Deviation",
  6,  "Merger briefed off the monitored channel",   "Deviation",
  8,  "@Elena faux pas (near-miss leak)",           "Near-miss",
  9,  "The Judge is installed",                     "Control",
  10, "SaltWind exposé #1",                         "External pressure",
  12, "SaltWind exposé #2",                         "External pressure",
  13, "Crisis exposé drops; stock down 8%",         "External pressure",
  17, "Legal and the Judge go offline",             "Control failure",
  18, "Judge approves disclosure with guardrails",  "Guardrail bent",
  19, "Judge issues an unenforceable warning",      "Guardrail bent",
  21, "EMBARGO BROKEN: merger confirmed publicly",  "Breach",
  22, "Embargo officially lifts",                   "Resolution"
)

kind_levels <- c("External pressure", "Deviation", "Near-miss", "Control",
                 "Control failure", "Guardrail bent", "Breach", "Resolution")
kind_cols <- c("External pressure" = "#9aa7b5", "Deviation" = "#fdae61",
               "Near-miss" = "#f16913", "Control" = "#4575b4",
               "Control failure" = "#6a51a3", "Guardrail bent" = "#fb6a4a",
               "Breach" = "#cb181d", "Resolution" = "#41ab5d")

events_plot <- events |>
  left_join(environment |> select(round, stress, hour, narrative), by = "round") |>
  mutate(
    kind = factor(kind, levels = kind_levels),
    narr = narrative |> str_replace_all("\\s*[\u2013\u2014]\\s*", ", ") |>
                        str_replace_all("'", ""),
    tip  = paste0("<b>", str_sub(hour, 6, 16), "</b><br>", narr)
  )

key_labels <- events_plot |> filter(round %in% c(3, 8, 9, 17, 21))

p_timeline <- ggplot(environment, aes(round, stress)) +
  annotate("rect", xmin = -0.5, xmax = 12.5, ymin = 0.5, ymax = 5.6,
           fill = "grey96") +
  annotate("rect", xmin = 12.5, xmax = 22.5, ymin = 0.5, ymax = 5.6,
           fill = "#fdecea") +
  geom_step(direction = "mid", colour = "grey55", linewidth = 0.8) +
  geom_vline(xintercept = 21, linetype = "dashed", colour = "#cb181d") +
  geom_point_interactive(
    data = events_plot,
    aes(round, stress, fill = kind, tooltip = tip, data_id = round),
    shape = 21, size = 4.6, colour = "white", stroke = 0.8) +
  geom_text_repel(
    data = key_labels, aes(round, stress, label = label),
    size = 3, fontface = "bold", colour = "grey20",
    box.padding = 0.7, max.overlaps = Inf, seed = 1, min.segment.length = 0) +
  scale_fill_manual(values = kind_cols, name = "Type of moment") +
  scale_x_continuous(breaks = environment$round, labels = x_lab$lab) +
  scale_y_continuous(breaks = 1:5,
                     labels = c("calm", "cautious", "worried", "critical", "peak")) +
  labs(title = "Two weeks of pressure, one hour of breach",
       subtitle = "Market stress is the backbone. The coloured chain is the sequence of decisions.",
       x = NULL, y = "Market stress") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
        panel.grid.minor = element_blank(),
        legend.position = "bottom")

girafe(ggobj = p_timeline, width_svg = 9.5, height_svg = 6.5,
       options = list(
         opts_hover(css = "stroke:black;stroke-width:1.5px;"),
         opts_tooltip(css = "background:#222;color:#fff;padding:8px;border-radius:5px;font-size:12px;max-width:340px;")))
```

Reading this top to bottom, the shape of the failure is already visible. The market mood is calm for the first week, then worry sets in once the National Housing Policy Institute report lands, and it never recovers. The orange and red chain shows that the company's own resistance to leaking was being lowered in parallel: a back channel quietly opened, a near-miss happened, a monitor was bolted on in response, and then on crisis day that monitor first went dark and was then bent twice before the wall finally fell at 5:00 PM.

## Who was actually talking to whom

A sequence of events is only half the picture. The legal team also needs the relationships. I used the `responding_to` field to build a directed reply network, where an arrow runs from the agent who was replied to toward the agent who replied. I dropped self replies so the structure between agents is clear.

```{r}
#| label: fig-network
#| fig-cap: "The reply network across all 23 rounds. A dense senior triangle of Legal, Platform-Trust and Social-Manager carries the decision load. The two breach actors, Legal and Social-Manager, sit at its core; PR and the interns are peripheral."
edges <- messages |>
  filter(!is.na(responding_to)) |>
  select(responding_to, to = agent_label) |>
  inner_join(messages |> select(message_id, from = agent_label),
             by = c("responding_to" = "message_id")) |>
  filter(from != to) |>
  count(from, to, name = "weight")

nodes <- messages |>
  count(agent_label, name = "messages") |>
  rename(name = agent_label)

g <- tbl_graph(nodes = nodes, edges = edges, directed = TRUE)

set.seed(2026)
ggraph(g, layout = "stress") +
  geom_edge_fan(aes(width = weight, alpha = weight),
                colour = "grey60",
                arrow = arrow(length = unit(2.2, "mm"), type = "closed"),
                end_cap = circle(7, "mm"), start_cap = circle(7, "mm")) +
  geom_node_point(aes(size = messages, fill = name),
                  shape = 21, colour = "white", stroke = 1) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3.4, fontface = "bold") +
  scale_edge_width_continuous(range = c(0.3, 2.6), guide = "none") +
  scale_edge_alpha_continuous(range = c(0.15, 0.85), guide = "none") +
  scale_size_continuous(range = c(5, 16), name = "Messages sent") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(title = "The senior triangle did the deciding") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "bottom")
```

What jumps out is that the conversation is not evenly spread. Legal, Platform-Trust, and Social-Manager form a tightly connected triangle that carries almost all of the back and forth, while PR and the interns hover at the edges. That matters for the verdict, because the two agents who eventually broke the embargo, Legal and Social-Manager, are not rogue outliers. They sit at the dead centre of the decision making. The leak came from the core, not the fringe.

## What the Judge was doing the whole time

The brief asks me to highlight the system elements that let the post past the embargo, and the most important one is the Judge itself. So I traced its behaviour round by round: how many times it spoke, and what stance it took. The bars below count the Judge's messages each hour, coloured by the posture I read from its content.

```{r}
#| label: fig-judge
#| fig-cap: "The Judge across the timeline. It reviews diligently for the first few days, then falls silent through the crisis morning, goes fully offline at 1:00 PM, returns only to approve a disclosure and issue a warning it cannot enforce, and is absent entirely when the embargo actually breaks at 5:00 PM."
posture_levels <- c("Not yet present", "Active review", "Silent / absent",
                    "Offline", "Approves (guardrails)", "Warns (no power)")
posture_cols <- c("Not yet present" = "#e0e0e0", "Active review" = "#41ab5d",
                  "Silent / absent" = "#bdbdbd", "Offline" = "#252525",
                  "Approves (guardrails)" = "#fe9929", "Warns (no power)" = "#cb181d")

judge <- environment |>
  select(round, hour, judge_offline) |>
  left_join(messages |> filter(agent_label == "Judge-Agent") |>
              count(round, name = "msgs"), by = "round") |>
  mutate(
    msgs = replace_na(msgs, 0),
    posture = case_when(
      judge_offline                  ~ "Offline",
      round %in% 9:12                ~ "Active review",
      round == 18                    ~ "Approves (guardrails)",
      round == 19                    ~ "Warns (no power)",
      round >= 13 & msgs == 0        ~ "Silent / absent",
      TRUE                           ~ "Not yet present"
    ),
    posture = factor(posture, levels = posture_levels),
    # give the offline round a visible bar height so it is not invisible
    plot_h = if_else(judge_offline, 1.2, as.numeric(msgs))
  )

ggplot(judge, aes(round, plot_h, fill = posture)) +
  geom_col(width = 0.8, colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = 21, linetype = "dashed", colour = "#cb181d") +
  annotate("text", x = 17, y = 1.7, label = "OFFLINE", size = 2.8,
           fontface = "bold", colour = "#252525") +
  annotate("text", x = 21, y = 5.4, label = "breach\n5 PM", size = 2.8,
           colour = "#cb181d", hjust = 0.5) +
  scale_fill_manual(values = posture_cols, name = "Judge posture", drop = FALSE) +
  scale_x_continuous(breaks = environment$round, labels = x_lab$lab) +
  labs(title = "The monitor was present, then silent, then bent, then gone",
       x = NULL, y = "Messages from the Judge") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
        legend.position = "bottom")
```

This single chart reframes the whole "system breakdown" question for me. The Judge did not malfunction in the way that word usually implies. It was a careful reviewer for four solid days. The problem is the shape of its absence and compromise on the one day it mattered: silent through the morning, fully offline at the worst hour, and then, when it did return, it approved and warned rather than blocked. By the time the embargo actually broke, the Judge was not in the room at all.

::: callout-tip
## What let the post past the embargo

Three system elements, visible in the figures above, did the damage in combination. First, the existence of an unmonitored Shadow channel meant sensitive coordination could happen where the Judge never looked. Second, the Judge went offline at 1:00 PM on crisis day, removing the brake at the exact hour the pressure peaked. Third, the Judge's design only reviewed wording, so once the agents reframed the leak as a regulatory obligation it could approve each "defensible" step without ever blocking the cumulative outcome.
:::

# Task 2: typical behaviour versus the breach {#sec-task2}

The breach was a new behaviour. To show how new, I need a clear picture of what normal looked like first, and then hold the crisis day up against it. Two views do this cleanly.

## The posting fingerprint, or how the roles inverted

The single most damning chart in this whole exercise is also the simplest. I counted how many times each agent posted to the public (official, personal, or anonymous) during the calm baseline versus during crisis day, and drew the two as a dumbbell.

```{r}
#| label: fig-fingerprint
#| fig-cap: "Public posting by agent, baseline versus crisis day. The advisors who never posted in public (Legal, Platform-Trust) take over the public channels, while the agent whose job is public communication (PR) goes completely silent."
public_ch <- c("official_post", "personal_post", "anonymous_post")

fingerprint <- messages |>
  mutate(phase = if_else(round <= 12, "baseline", "crisis")) |>
  filter(channel %in% public_ch) |>
  count(agent_label, phase) |>
  pivot_wider(names_from = phase, values_from = n, values_fill = 0)

ggplot(fingerprint, aes(y = fct_reorder(agent_label, crisis))) +
  geom_segment(aes(x = baseline, xend = crisis, yend = agent_label),
               colour = "grey75", linewidth = 1.6) +
  geom_point(aes(x = baseline), colour = "#2c7fb8", size = 4.5) +
  geom_point(aes(x = crisis),   colour = "#cb181d", size = 4.5) +
  geom_text(aes(x = baseline, label = baseline), vjust = -1.3,
            colour = "#2c7fb8", size = 3.2) +
  geom_text(aes(x = crisis, label = crisis), vjust = -1.3,
            colour = "#cb181d", size = 3.2) +
  annotate("text", x = 1, y = 6.35, label = "Baseline (17 May to 4 Jun)",
           colour = "#2c7fb8", fontface = "bold", hjust = 0, size = 3.4) +
  annotate("text", x = 13, y = 6.35, label = "Crisis day (5 Jun)",
           colour = "#cb181d", fontface = "bold", hjust = 0, size = 3.4) +
  scale_x_continuous(limits = c(-0.5, 17), breaks = seq(0, 16, 4)) +
  labs(title = "The advisors took over the megaphone",
       x = "Number of public posts", y = NULL) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        plot.margin = margin(20, 20, 10, 10))
```

I find this almost cinematic. In normal times Legal and Platform-Trust never speak in public, not once across two weeks, exactly as you would expect from a legal advisor and a trust-and-safety lead. PR carries the official voice, and the interns post cheerful personal updates. On crisis day the roles flip. Legal jumps from zero public posts to sixteen, Platform-Trust from zero to five, and the agent whose entire purpose is public communication, PR, drops from six to absolute silence. When the people who are supposed to hold the line start running the public account, and the people whose job is the public account disappear, that is not business as usual. That is the anomaly, drawn in two colours.

## When everyone got loud at once

The dumbbell counts public posts only. Zooming out to total message volume tells a complementary story about tempo. The heatmap below shows how many messages each agent sent in every hour, with darker cells meaning more talk. Hover any cell for the exact count and that hour's headline.

```{r}
#| label: fig-heatmap
#| fig-cap: "Message volume by agent and hour. The calm baseline is a faint, even wash. Crisis day is a dark wall of traffic concentrated in the senior triangle, while PR and the interns stay pale. The intensity is itself a signal of decisions being made at speed."
heat <- messages |>
  count(agent_label, round, name = "msgs") |>
  complete(agent_label, round = 0:22, fill = list(msgs = 0)) |>
  left_join(environment |> select(round, hour, headline), by = "round") |>
  mutate(
    agent_label = fct_reorder(agent_label, msgs, .fun = sum),
    when = if_else(round <= 12, str_sub(hour, 6, 10), str_sub(hour, 12, 16)),
    tip  = paste0("<b>", agent_label, "</b><br>", when,
                  "<br>messages: ", msgs,
                  "<br>", str_replace_all(headline %||% "", "'", ""))
  )

p_heat <- ggplot(heat, aes(round, agent_label, fill = msgs)) +
  geom_tile_interactive(aes(tooltip = tip, data_id = paste(agent_label, round)),
                        colour = "white", linewidth = 0.4) +
  geom_vline(xintercept = 12.5, colour = "grey30", linewidth = 0.4) +
  geom_vline(xintercept = 21, linetype = "dashed", colour = "#cb181d") +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b", name = "Messages") +
  scale_x_continuous(breaks = environment$round, labels = x_lab$lab,
                     expand = c(0, 0)) +
  labs(title = "A pale baseline, then a wall of crisis-day traffic",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
        panel.grid = element_blank())

girafe(ggobj = p_heat, width_svg = 9.5, height_svg = 4.8,
       options = list(
         opts_hover(css = "stroke:black;stroke-width:1px;"),
         opts_tooltip(css = "background:#222;color:#fff;padding:8px;border-radius:5px;font-size:12px;max-width:320px;")))
```

The vertical line marks the switch from daily to hourly rounds, so some of the darkening is just finer time slicing. But the concentration is real. Almost all of the crisis-day traffic piles into Legal, Social-Manager, and Platform-Trust, the same triangle from the network. PR and the interns barely register once the pressure is on. The volume is not just noise. It is the visible texture of a small group making rapid, high-stakes calls with very little friction between them.

## Where the conversation went to hide

The second view tracks the centre of gravity of communication. I plotted each channel's share of all messages, round by round, ordering the channels from the most accountable at the bottom to the most deniable at the top.

```{r}
#| label: fig-channels
#| fig-cap: "Channel composition over time. In calm weeks the conversation lives in the monitored team room and the sanctioned official feed. On crisis day it migrates upward into the Shadow channel and, for the first time ever, anonymous public posts."
chan_levels <- c("comms_huddle", "official_post", "one_on_one_chat",
                 "personal_post", "side_huddle", "anonymous_post")
chan_cols <- c("comms_huddle" = "#08519c", "official_post" = "#3182bd",
               "one_on_one_chat" = "#9ecae1", "personal_post" = "#fdae6b",
               "side_huddle" = "#e6550d", "anonymous_post" = "#a50f15")

chan_share <- messages |>
  count(round, channel) |>
  group_by(round) |>
  mutate(share = n / sum(n)) |>
  ungroup() |>
  mutate(channel = factor(channel, levels = chan_levels))

ggplot(chan_share, aes(round, share, fill = channel)) +
  geom_area(colour = "white", linewidth = 0.15) +
  geom_vline(xintercept = 12.5, linetype = "dashed", colour = "grey30") +
  geom_vline(xintercept = 21, linetype = "dashed", colour = "#cb181d") +
  annotate("text", x = 6, y = 1.03, label = "monitored / sanctioned (cool)",
           size = 3, colour = "#08519c", fontface = "bold") +
  annotate("text", x = 18.5, y = 1.03, label = "shadow / anonymous (warm)",
           size = 3, colour = "#a50f15", fontface = "bold") +
  scale_fill_manual(values = chan_cols, name = "Channel") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1.06),
                     breaks = seq(0, 1, 0.25)) +
  scale_x_continuous(breaks = environment$round, labels = x_lab$lab) +
  labs(title = "The conversation drifted toward channels nobody could see",
       x = NULL, y = "Share of messages in the hour") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
        legend.position = "bottom")
```

The warm colours are almost absent for the first ten rounds. The Shadow channel flickers on once around 22 May and then goes quiet again. Then crisis day arrives and the warm band swells. The Shadow channel becomes a permanent fixture, and the deepest red, the anonymous public post, appears for the very first time in the entire fortnight. A channel that did not exist in the team's behaviour suddenly carries the most sensitive content of the day. The escape route was not just used on crisis day. It was, in effect, invented on crisis day.

# Task 3: were there leading indicators {#sec-task3}

The legal team will want to know whether anyone could have seen this coming. They could have. The signals were there, they simply did not look like a crisis at the time.

## The tone of private reasoning shifted before the breach

This is the finding I am most pleased with, and it comes entirely from the `internal_state` field. I counted, for each round, how many private thoughts were tagged as deliberating, reacting, or rationalizing, and stacked them.

```{r}
#| label: fig-tone
#| fig-cap: "How the agents reasoned privately, round by round. Calm deliberation is the only mode for two weeks. On crisis day it gives way first to raw reaction, and then, precisely as the embargo is being pushed past, to self-justification."
mode_levels <- c("deliberating", "reacting", "rationalizing")
mode_cols <- c("deliberating" = "#3182bd", "reacting" = "#fe9929",
               "rationalizing" = "#cb181d")

tone <- messages |>
  select(round, deliberating, reacting, rationalizing) |>
  pivot_longer(c(deliberating, reacting, rationalizing),
               names_to = "mode", values_to = "text") |>
  filter(!is.na(text)) |>
  count(round, mode) |>
  mutate(mode = factor(mode, levels = mode_levels))

ggplot(tone, aes(round, n, fill = mode)) +
  geom_col(width = 0.8, colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = 12.5, linetype = "dashed", colour = "grey30") +
  geom_vline(xintercept = 21, linetype = "dashed", colour = "#cb181d") +
  annotate("text", x = 6, y = 5.6, label = "deliberation only",
           size = 3, colour = "#3182bd", fontface = "bold") +
  annotate("text", x = 20.2, y = 3.6, label = "rationalizing\nappears",
           size = 3, colour = "#cb181d", fontface = "bold") +
  scale_fill_manual(values = mode_cols, name = "Private reasoning mode") +
  scale_x_continuous(breaks = environment$round, labels = x_lab$lab) +
  labs(title = "Deliberation, then reaction, then rationalization",
       x = NULL, y = "Private thoughts logged") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
        legend.position = "bottom")
```

For the entire pre-crisis fortnight, every single logged thought is blue. The agents only ever deliberate. They weigh, they hedge, they qualify. On crisis day the colour changes. The mornings of 5 June are dominated by amber reaction, the agents responding to a stock in freefall and a hostile press. And then, in the late afternoon rounds when the embargo is actually being argued away, red appears for the first time. Rationalizing thoughts show up at exactly 2:00 PM and grow through the breach. This is the fingerprint of my whole thesis. The mode of thinking flips from "what is the right thing to do" to "here is why the thing I want to do is acceptable", and it flips in the same hours that the wall comes down.

## The vocabulary of justification

If the mode of reasoning changed, did the words change with it? I ran a small text-mining pass over the internal-state fields, splitting them into the calm fortnight and crisis day, and pulled out the most distinctive words in each. The two panels below are, in a sense, the dictionary the agents reached for in each period.

```{r}
#| label: fig-lexicon
#| fig-cap: "The language of private reasoning, before and during the crisis. Calm deliberation is about Ajay, the board, what is right, and risk. Crisis reasoning is about the embargo, the merger, silence, the narrative, and the fact that things are 'already' public. The escape-hatch vocabulary appears exactly when the embargo is being argued away."
internal_text <- messages |>
  mutate(phase = if_else(round <= 12,
                         "Pre-crisis reasoning", "Crisis-day reasoning")) |>
  select(phase, reacting, rationalizing, deliberating) |>
  pivot_longer(c(reacting, rationalizing, deliberating), values_to = "text") |>
  filter(!is.na(text)) |>
  select(phase, text)

lex <- internal_text |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(str_length(word) >= 4, !str_detect(word, "[0-9]")) |>
  count(phase, word, sort = TRUE) |>
  group_by(phase) |>
  slice_max(n, n = 12, with_ties = FALSE) |>
  ungroup() |>
  mutate(phase = factor(phase,
                        levels = c("Pre-crisis reasoning", "Crisis-day reasoning")))

ggplot(lex, aes(n, reorder_within(word, n, phase), fill = phase)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  facet_wrap(~phase, scales = "free") +
  scale_y_reordered() +
  scale_fill_manual(values = c("Pre-crisis reasoning" = "#3182bd",
                               "Crisis-day reasoning" = "#cb181d")) +
  labs(title = "From 'what is right' to 'why this is allowed'",
       x = "Word frequency in internal states", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        panel.grid.major.y = element_blank())
```

I do not want to over-read a small corpus, but the shift is hard to miss. Before the crisis the agents privately think about Ajay, the board, the story, and what the right call is. During the crisis the most frequent words become embargo, merger, silence, narrative, and especially "already", as in the market already knows and SaltWind already published. That last word is the hinge of the entire rationalization. Once "the information is already out" becomes the dominant frame, confirming it stops feeling like a breach and starts feeling like catching up. The words did not just change. They changed into the exact language used to excuse the leak.

## The near-miss ledger

There were three earlier moments where behaviour departed from what was expected. I have set them out below with what was observed and, crucially, why each one was waved through.

| When | What was observed | Why nothing real happened |
|----|----|----|
| 22 May (R3) | The Shadow side channel was opened for the first time, and the senior team began coordinating where the monitor could not see. | It was framed as legitimate crisis coordination after the NHPI report. A private back channel looked like prudence, not risk. |
| 25 May (R6) | The merger itself was briefed to the senior team inside the Shadow channel, deliberately off the monitored room. | Nobody flagged that the most sensitive fact in the company had been routed around oversight by design, because the routing felt natural. |
| 29 May (R8) | The @Elena faux pas: a personal post tagged CivicLoom's CEO with "big things coming", and a CivicLoom account liked it before deletion. A genuine near-leak. | This one *did* trigger a response. The Judge was installed. But the response treated the symptom, one careless post, and never addressed the structural risk that pressured agents would rationalize disclosure. |

The pattern is consistent. Each deviation was individually explainable, so each was individually excused. The @Elena incident is the telling one. It was the system working as intended, a near-miss caught and answered, yet the answer was to add a wording monitor rather than to ask why the team kept drifting toward the edge. The monitor that was installed in response to a near-leak is the very monitor that was later argued into approving the real one.

::: callout-important
## Why the warning signs were never enough

The Judge was reactive, it reviewed phrasing rather than intent, and it could be reasoned with. None of the leading indicators looked dangerous in isolation, and the one control that did exist was structurally blind to a slow, collective, well-argued drift. The signals were not missed because they were hidden. They were missed because each one had a reasonable explanation, and reasonable explanations were exactly the weapon used to break the embargo.
:::

# My verdict {#sec-verdict}

So, deliberate leak or system breakdown? Having lived in this data, I do not think either label is honest on its own.

It was not a clean technical breakdown. The Judge functioned correctly for almost the entire period. It reviewed content, softened claims, and held items back during the calm weeks. Its only hard failure was a one-hour outage at 1:00 PM on crisis day. A system that worked for twenty-two of twenty-three rounds did not "crash".

But it was also not a premeditated conspiracy. Nowhere in 912 messages does an agent say, in effect, let us leak the merger early for advantage. There is no plot. What there is, in the private internal states, is a cascade of justification under unbearable pressure. The stock was collapsing, a loan covenant had broken, and the acquiring party was threatening to walk. In that vice, the agents reframed the leak step by step into things that each sounded reasonable: a regulatory obligation to a congressional inquiry, a securities-law duty not to stay silent, a contractual clause permitting early release by mutual consent, and finally the claim that since a newspaper had already published the story, confirming it was no longer a breach at all.

The Judge did not stop any of this, and the reason is the heart of the matter. It was built to police wording, so it could be argued into approving each individually defensible sentence while never being equipped to block the cumulative outcome those sentences added up to. At 2:00 PM it approved a disclosure "with guardrails". At 3:00 PM it issued a warning it had no power to enforce. By 5:00 PM it had been routed around entirely.

That is why I keep coming back to the same sentence. The embargo was not hacked and it did not crash. It was talked past. The failure was neither malice nor malfunction but rationalization, the most human failure of all, executed at machine speed by the very agents the system was built to watch. For CivicLoom's purposes that distinction matters enormously, because it means the fix is not a better firewall. It is a monitor that can reason about cumulative intent, and a culture that does not let "this is technically defensible" become the last thing anyone says before the wall comes down.

# Reflection and takeaways {#sec-reflection}

A few things stuck with me from building this.

The most valuable variable was the one that was easiest to ignore. Anyone can chart who posted what. Almost nobody, I suspect, will chart how the agents privately justified themselves, and yet the tone-shift in @fig-tone is the clearest single piece of evidence in the whole exercise. The lesson I am taking forward is to hunt for the field that records reasoning, not just the field that records action.

Cleaning genuinely changed the conclusion. The contaminated price field in @sec-wrangle would have led me to draw a stock that looked already dead, when the real story was a revenue figure tripping a covenant. Good wrangling is not housekeeping before the analysis. Sometimes it is the analysis.

::: callout-note
## My takeaway

If I had to compress this into one line for the legal team: nobody broke the embargo, everybody reasoned it away. The next version of this company does not need a stronger lock. It needs a monitor that can see a chain of small, defensible steps for what they add up to.
:::
