---
title: "Hands-on Exercise 1: A Layered Grammar of Graphics"
author: "Meng Tingyu"
date: "2026-04-15"
date-modified: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-tools: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## 1. Overview

This exercise explores the foundational principles of **ggplot2**, an R package for declaratively creating data-driven graphics based on Leland Wilkinson's *Grammar of Graphics*. Unlike Base R graphics which relies on a series of function calls, ggplot2 adopts a structured **layered approach** that makes visualisations more readable, reproducible, and aesthetically consistent.

The key idea behind ggplot2 can be summarised as:

> *"The transferable skills from ggplot2 are not the idiosyncrasies of plotting syntax, but a powerful way of thinking about visualisation, as a way of mapping between variables and the visual properties of geometric objects that you can perceive."* — Hadley Wickham

---

## 2. Getting Started

### 2.1 Loading Required Libraries

```{r}
pacman::p_load(tidyverse)
```

### 2.2 Importing the Dataset

```{r}
exam_data <- read_csv("data/Exam_data.csv")
glimpse(exam_data)
```

The dataset contains year-end examination grades of a cohort of Primary 3 students from a local school. It consists of **seven attributes**:

| Type | Variables |
|------|-----------|
| Categorical | ID, CLASS, GENDER, RACE |
| Continuous | MATHS, ENGLISH, SCIENCE |

---

## 3. The Grammar of Graphics

ggplot2 builds every plot from **seven layered components**. Understanding these components is essential to constructing meaningful visualisations.

| Component | Role |
|-----------|------|
| **Data** | The dataset being plotted |
| **Aesthetics** | Maps data variables to visual properties (position, colour, size, shape) |
| **Geometrics** | The visual marks used to represent data (points, bars, lines) |
| **Facets** | Splits data into subplots by a categorical variable |
| **Statistics** | Applies statistical transformations (mean, confidence intervals) |
| **Coordinates** | Defines the plane on which data is mapped |
| **Themes** | Controls non-data elements such as background and fonts |

A plot is built by stacking these layers using the `+` operator.

---

## 4. Base R Graphics vs ggplot2

To appreciate the advantages of ggplot2, it is useful to compare it with Base R graphics using the same dataset.

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Base R: hist()"
#|   - "ggplot2: geom_histogram()"

# Base R
hist(exam_data$MATHS,
     main = "Distribution of Maths Scores",
     xlab = "Maths Score",
     col = "grey80")

# ggplot2
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 fill = "steelblue",
                 color = "white") +
  labs(title = "Distribution of Maths Scores",
       x = "Maths Score",
       y = "Count") +
  theme_minimal()
```

While Base R offers a quick way to produce plots, ggplot2 provides a more systematic and extensible framework, especially when dealing with complex multi-layered visualisations.

---

## 5. Essential Geometric Objects

Geometric objects (geoms) define **what type of mark** is drawn on the plot. Below are the key geoms introduced in this exercise.

### 5.1 Bar Chart — `geom_bar()`

Used to visualise the frequency of a categorical variable.

```{r}
ggplot(exam_data, aes(x = RACE)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Number of Students by Race",
       x = "Race", y = "Count") +
  theme_minimal()
```

### 5.2 Dot Plot — `geom_dotplot()`

Each dot represents one observation. Useful for small datasets to see individual data points.

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot(binwidth = 2.5,
               dotsize = 0.5,
               fill = "steelblue") +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Dot Plot of Maths Scores",
       x = "Maths Score") +
  theme_minimal()
```

::: callout-note
The y-axis is suppressed here as it does not carry meaningful information in a dot plot.
:::

### 5.3 Histogram — `geom_histogram()`

Used to visualise the distribution of a continuous variable.

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Default histogram"
#|   - "Coloured by gender"

ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 fill = "steelblue",
                 color = "white") +
  labs(title = "Maths Score Distribution",
       x = "Maths Score", y = "Count") +
  theme_minimal()

ggplot(exam_data, aes(x = MATHS, fill = GENDER)) +
  geom_histogram(bins = 20,
                 color = "grey30") +
  labs(title = "Maths Score Distribution by Gender",
       x = "Maths Score", y = "Count") +
  theme_minimal()
```

### 5.4 Kernel Density Plot — `geom_density()`

A smoothed version of the histogram, useful for visualising the underlying distribution.

```{r}
ggplot(exam_data, aes(x = MATHS, colour = GENDER)) +
  geom_density(linewidth = 0.8) +
  labs(title = "Kernel Density Estimate of Maths Scores by Gender",
       x = "Maths Score", y = "Density") +
  theme_minimal()
```

### 5.5 Boxplot — `geom_boxplot()`

Displays the five-number summary (minimum, Q1, median, Q3, maximum) and highlights outliers.

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Standard boxplot"
#|   - "Notched boxplot"

ggplot(exam_data, aes(x = GENDER, y = MATHS)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(title = "Maths Scores by Gender",
       x = "Gender", y = "Maths Score") +
  theme_minimal()

ggplot(exam_data, aes(x = GENDER, y = MATHS)) +
  geom_boxplot(notch = TRUE,
               fill = "steelblue", alpha = 0.6) +
  labs(title = "Notched Boxplot of Maths Scores by Gender",
       x = "Gender", y = "Maths Score") +
  theme_minimal()
```

::: callout-tip
If the notches of two boxplots do not overlap, there is strong evidence that their medians differ.
:::

### 5.6 Violin Plot — `geom_violin()`

Combines a boxplot with a kernel density estimate, showing both summary statistics and the full distribution shape.

```{r}
ggplot(exam_data, aes(x = GENDER, y = MATHS)) +
  geom_violin(fill = "steelblue", alpha = 0.6) +
  labs(title = "Violin Plot of Maths Scores by Gender",
       x = "Gender", y = "Maths Score") +
  theme_minimal()
```

### 5.7 Scatterplot — `geom_point()`

Used to visualise the relationship between two continuous variables.

```{r}
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.6, colour = "steelblue") +
  labs(title = "Maths vs English Scores",
       x = "Maths Score", y = "English Score") +
  theme_minimal()
```

### 5.8 Combining Geoms

Multiple geoms can be layered on the same plot using `+`.

```{r}
ggplot(exam_data, aes(x = GENDER, y = MATHS)) +
  geom_boxplot() +
  geom_point(position = "jitter",
             size = 0.5,
             alpha = 0.4,
             colour = "steelblue") +
  labs(title = "Boxplot with Jittered Data Points",
       x = "Gender", y = "Maths Score") +
  theme_minimal()
```

---

## 6. Statistical Transformations

Statistical functions (`stat_*`) allow us to overlay computed summaries on top of raw data.

### 6.1 Adding Mean to Boxplot

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Using stat_summary()"
#|   - "Using geom_point(stat = 'summary')"

ggplot(exam_data, aes(x = GENDER, y = MATHS)) +
  geom_boxplot() +
  stat_summary(geom = "point",
               fun = "mean",
               colour = "red",
               size = 4) +
  labs(title = "stat_summary() method",
       x = "Gender", y = "Maths Score") +
  theme_minimal()

ggplot(exam_data, aes(x = GENDER, y = MATHS)) +
  geom_boxplot() +
  geom_point(stat = "summary",
             fun = "mean",
             colour = "red",
             size = 4) +
  labs(title = "geom_point(stat = 'summary') method",
       x = "Gender", y = "Maths Score") +
  theme_minimal()
```

### 6.2 Best Fit Curve — `geom_smooth()`

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "LOESS smoothing (default)"
#|   - "Linear model"

ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.5) +
  geom_smooth(linewidth = 0.8) +
  labs(title = "LOESS Smoothing",
       x = "Maths Score", y = "English Score") +
  theme_minimal()

ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = lm, linewidth = 0.8) +
  labs(title = "Linear Model",
       x = "Maths Score", y = "English Score") +
  theme_minimal()
```

---

## 7. Facets

Facets generate **small multiples** — the same plot repeated across subsets of data. This is particularly useful for comparing distributions across groups.

### 7.1 `facet_wrap()`

Wraps panels into a 2D grid, ideal for a single grouping variable.

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 fill = "steelblue",
                 color = "white") +
  facet_wrap(~ CLASS) +
  labs(title = "Maths Score Distribution by Class",
       x = "Maths Score", y = "Count") +
  theme_minimal()
```

### 7.2 `facet_grid()`

Forms a matrix of panels defined by two grouping variables.

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 fill = "steelblue",
                 color = "white") +
  facet_grid(~ CLASS) +
  labs(title = "Maths Score Distribution by Class (facet_grid)",
       x = "Maths Score", y = "Count") +
  theme_minimal()
```

---

## 8. Coordinates

Coordinate systems control how data is mapped onto the 2D plane of the plot.

### 8.1 `coord_flip()`

Flips the x and y axes — useful for horizontal bar charts with long category labels.

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Vertical bar chart (default)"
#|   - "Horizontal bar chart (coord_flip)"

ggplot(exam_data, aes(x = RACE)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Default: Vertical",
       x = "Race", y = "Count") +
  theme_minimal()

ggplot(exam_data, aes(x = RACE)) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  labs(title = "Flipped: Horizontal",
       x = "Race", y = "Count") +
  theme_minimal()
```

### 8.2 `coord_cartesian()` — Fixing Axis Ranges

Setting equal axis ranges ensures scatterplots are not visually misleading.

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "Default axis range"
#|   - "Fixed axis range (0–100)"

ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = lm, linewidth = 0.5) +
  labs(title = "Default Axis Range",
       x = "Maths Score", y = "English Score") +
  theme_minimal()

ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = lm, linewidth = 0.5) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  labs(title = "Fixed Axis Range (0–100)",
       x = "Maths Score", y = "English Score") +
  theme_minimal()
```

---

## 9. Themes

Themes control the **non-data elements** of a plot: background colour, grid lines, font sizes, and axis labels.

```{r}
#| layout-ncol: 3
#| fig-cap:
#|   - "theme_gray() — default"
#|   - "theme_classic()"
#|   - "theme_minimal()"

ggplot(exam_data, aes(x = RACE)) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  labs(title = "theme_gray()") +
  theme_gray()

ggplot(exam_data, aes(x = RACE)) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  labs(title = "theme_classic()") +
  theme_classic()

ggplot(exam_data, aes(x = RACE)) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  labs(title = "theme_minimal()") +
  theme_minimal()
```

::: callout-tip
`theme_minimal()` is generally preferred in academic and data journalism contexts as it reduces visual clutter and keeps the focus on the data.
:::

---

## 10. Key Takeaways

This exercise demonstrates that ggplot2's layered grammar provides a principled and flexible framework for building statistical graphics. The key insights are:

- **Start with data and aesthetics** — define what variables map to which visual properties
- **Choose the right geom** — different geoms suit different data types and analytical goals
- **Layer deliberately** — combining geoms (e.g. boxplot + jitter) reveals more information than either alone
- **Use facets for comparison** — small multiples are more effective than overloaded single plots
- **Apply statistical summaries** — overlaying means or trend lines adds analytical value
- **Fix axis ranges** — always set equal axis limits when comparing two variables on the same scale
- **Choose themes purposefully** — `theme_minimal()` or `theme_classic()` are generally preferred for clean, publication-ready visuals

---

## References

- Hadley Wickham (2010). "A layered grammar of graphics." *Journal of Computational and Graphical Statistics*, 19(1), 3–28.
- Leland Wilkinson (1999). *Grammar of Graphics*. Springer.
- Garrett Grolemund & Hadley Wickham (2023). *R for Data Science (2e)*. O'Reilly.

