---
title: "Hands-on Exercise 2: Beyond ggplot2 Fundamentals"
author: "Meng Tingyu"
date: "2026-04-23"
date-modified: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-tools: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## 1. Overview

This exercise builds on the foundations of ggplot2 by introducing four powerful extension packages that enhance the quality and communicability of statistical graphics:

| Package | Purpose |
|---------|---------|
| **ggrepel** | Automatically repositions overlapping text labels on plots |
| **ggthemes** | Provides additional publication-ready themes (e.g. The Economist, WSJ) |
| **hrbrthemes** | Provides typography-focused themes optimised for readability |
| **patchwork** | Combines multiple ggplot2 plots into a single composite figure |

By the end of this exercise, you will be able to produce clean, professional-quality visualisations suitable for academic and industry reporting.

---

## 2. Getting Started

### 2.1 Loading Required Libraries

```{r}
pacman::p_load(ggrepel, patchwork,
               ggthemes, hrbrthemes,
               tidyverse)
```

### 2.2 Importing the Dataset

```{r}
exam_data <- read_csv("data/Exam_data.csv")
glimpse(exam_data)
```

The dataset contains year-end examination grades of Primary 3 students, with four categorical variables (ID, CLASS, GENDER, RACE) and three continuous variables (MATHS, ENGLISH, SCIENCE).

---

## 3. Beyond ggplot2 Annotation: ggrepel

When plotting a scatterplot with many data points, adding labels using `geom_label()` or `geom_text()` often results in **overlapping text** that is difficult to read.

### 3.1 The Problem: Overlapping Labels

```{r}
ggplot(exam_data,
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5) +
  geom_label(aes(label = ID),
             hjust = 0.5,
             vjust = -0.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  labs(title = "English scores versus Maths scores for Primary 3",
       x = "Maths Score", y = "English Score") +
  theme_minimal()
```


### 3.2 The Solution: ggrepel

`ggrepel` solves this by automatically adjusting label positions to avoid overlap. Simply replace:

- `geom_label()` → `geom_label_repel()`
- `geom_text()` → `geom_text_repel()`

```{r}
ggplot(exam_data,
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5) +
  geom_label_repel(aes(label = ID),
                   fontface = "bold",
                   max.overlaps = 20) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  labs(title = "English scores versus Maths scores for Primary 3",
       x = "Maths Score", y = "English Score") +
  theme_minimal()
```


---

## 4. Beyond ggplot2 Themes

### 4.1 Built-in ggplot2 Themes

ggplot2 comes with eight built-in themes. The most commonly used are:

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "theme_gray() — default"
#|   - "theme_classic()"

ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  labs(title = "Distribution of Maths Scores") +
  theme_gray()

ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  labs(title = "Distribution of Maths Scores") +
  theme_classic()
```

### 4.2 ggthemes Package

`ggthemes` provides themes that replicate the visual style of well-known publications and tools, such as The Economist, FiveThirtyEight, The Wall Street Journal, and Stata.

```{r}
#| layout-ncol: 2
#| fig-cap:
#|   - "theme_economist()"
#|   - "theme_wsj()"

ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  labs(title = "Distribution of Maths Scores") +
  theme_economist()

ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  labs(title = "Distribution of Maths Scores") +
  theme_wsj()
```


### 4.3 hrbrthemes Package

`hrbrthemes` focuses on **typographic clarity** — it optimises font sizes, label placement, and grid lines for readability in a production workflow.

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  labs(title = "Distribution of Maths Scores") +
  theme_ipsum()
```

We can further customise the typography using additional arguments:

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  labs(title = "Distribution of Maths Scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```


---

## 5. Beyond Single Graph: patchwork

In data storytelling, a single chart is rarely sufficient. `patchwork` allows us to **combine multiple ggplot2 plots** into one cohesive composite figure using simple, intuitive operators.

### 5.1 Creating the Individual Plots

We first create three plots separately and store them as objects:

```{r}
p1 <- ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  coord_cartesian(xlim = c(0, 100)) +
  labs(title = "Distribution of Maths Scores") +
  theme_minimal()

p2 <- ggplot(exam_data, aes(x = ENGLISH)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  coord_cartesian(xlim = c(0, 100)) +
  labs(title = "Distribution of English Scores") +
  theme_minimal()

p3 <- ggplot(exam_data,
             aes(x = MATHS, y = ENGLISH)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = lm, linewidth = 0.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  labs(title = "English vs Maths Scores",
       x = "Maths Score", y = "English Score") +
  theme_minimal()
```

### 5.2 Two-Column Layout: `+`

The `+` operator places two plots **side by side**:

```{r}
p1 + p2
```

### 5.3 Three-Plot Composite Layout

Using a combination of `/` (stack vertically) and `|` (place side by side):

```{r}
(p1 / p2) | p3
```


### 5.4 Adding Tags to Subplots

`plot_annotation()` automatically labels each subplot, making it easy to reference them in text:

```{r}
((p1 / p2) | p3) +
  plot_annotation(tag_levels = "I")
```


### 5.5 Inset Plot

`inset_element()` allows you to embed a smaller plot **inside** a larger one, useful for showing a zoomed-in view or a supplementary chart:

```{r}
p3 + inset_element(p2,
                   left = 0.02,
                   bottom = 0.7,
                   right = 0.5,
                   top = 1)
```


### 5.6 Combining patchwork with ggthemes

The `&` operator applies a theme to **all plots** in a patchwork composite simultaneously:

```{r}
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

::: callout-note
Notice the difference between `+` and `&` in patchwork:

- `+` adds elements to the **last plot** only
- `&` applies the element to **all plots** in the composite
:::

---

## 6. Key Takeaways

This exercise let me learn how ggplot2 extensions can significantly elevate the quality of statistical graphics:

- **ggrepel** solves the overlapping label problem automatically — always prefer `geom_label_repel()` over `geom_label()` when labelling individual data points
- **ggthemes** provides ready-made publication themes that immediately give your plots a professional appearance
- **hrbrthemes** is particularly useful when typography and readability are a priority
- **patchwork** makes it easy to tell a multi-panel visual story with minimal code — use `/` to stack, `|` to place side by side, and `&` to apply themes globally
- Composite figures with `plot_annotation(tag_levels = "I")` are especially important in academic writing where subplots need to be referenced in text

---

