---
title: "Take-home Exercise 1"
subtitle: "Visual Analytics for Insurance Portfolio Profitability & Volatility"
author: "Meng Tingyu"
date: "`r Sys.Date()`"
date-modified: "`r Sys.Date()`"
format:
  html:
    code-fold: show
    code-tools: false
    code-summary: "Show code"
    code-line-numbers: false
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    fig-width: 10
    fig-height: 6
    fig-align: center
    fig-cap-location: bottom
    tbl-cap-location: top
    theme: [cosmo, custom.scss]
    css: custom.css
    smooth-scroll: true
    embed-resources: false
execute:
  warning: false
  message: false
  freeze: auto
  cache: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 96,
  out.width = "100%"
)
options(scipen = 999, digits = 4)
```

# Overview {.unnumbered}

> *"Insurers often carry underperforming portfolios where the root causes of poor profitability and high volatility are unknown. Leveraging visual analytics can potentially help investigate trends in underlying loss drivers."*

This study applies a **visual-analytics workflow** to a real-world Spanish motor-insurance portfolio (354,140 records over 2022–2024) to discover the drivers of **profitability** and **volatility**. Rather than producing yet another dashboard, the analysis combines univariate exploration, bivariate hypothesis-testing, and interactive cross-filtering to surface non-obvious portfolio dynamics that direct downstream underwriting decisions.

::: callout-note
## Working definitions

**Profitability** is operationalised as the **Loss Ratio** (`LR = total_incurred / total_premium`). Values \< 1 indicate the premium covers claims; \> 1 indicates underwriting loss. We complement LR with **Pure Premium** (loss per unit exposure), **Claim Frequency** (claims per unit exposure) and **Severity** (cost per claim).

**Volatility** is operationalised through three lenses:

\(i\) the **coefficient of variation** of policy-level LR within each segment;

\(ii\) **tail concentration**: the share of total losses generated by the largest claims (Lorenz/Gini);

\(iii\) the **dispersion of severity** within segments.
:::

# Analytical Workflow

```{mermaid}
%%| fig-cap: "Analytical workflow from raw policy data to actionable insight."
flowchart LR
  A[Raw CSV<br/>354k rows · 47 cols<br/>; delimiter] --> B[Data Wrangling<br/>tidyverse]
  B --> C[Analytical Sandbox<br/>+ derived KPIs]
  C --> D1[Univariate EDA<br/>Numerical]
  C --> D2[Univariate EDA<br/>Categorical]
  D1 --> E[Bivariate Analysis<br/>+ Statistical Validation]
  D2 --> E
  E --> F1[Profitability Drivers]
  E --> F2[Volatility &amp; Tail Risk]
  F1 --> G[Interactive<br/>Cross-filtering<br/>plotly · ggiraph]
  F2 --> G
  G --> H[Underwriting<br/>Insight]
  style A fill:#f5f5f5,stroke:#333333,stroke-width:1px
  style C fill:#fafafa,stroke:#222222,stroke-width:1.5px
  style G fill:#ffffff,stroke:#000000,stroke-width:2px
  style H fill:#222222,stroke:#000000,color:#ffffff
```

# Data & Methodology

## Data source

The dataset is a real-world motor-vehicle insurance portfolio published by a Spanish non-life insurer on [Mendeley Data](https://data.mendeley.com/). It contains **354,140 policy-year records** for **185,678 unique policies** across calendar years **2022, 2023 and 2024**, with 47 variables grouped into:

-   **Policy & insured** (8 vars): `policy_type`, `policy_status`, `business_type`, `payment_frequency`, `bonus_score`, etc.
-   **Driver & vehicle** (10 vars): `driver_age`, `vehicle_age`, `fuel_type`, `vehicle_value`, `vehicle_brand`, `municipality_type`, `circulation_area`, …
-   **Premiums** (8 vars): `total_premium` and 7 coverage-level premiums.
-   **Claim counts** (10 vars): `total_claims` plus per-coverage counts.
-   **Incurred losses** (10 vars): `total_incurred` plus per-coverage losses.
-   **Exposure** (2 vars): policy-year exposure used to annualise frequencies.

::: callout-warning
## Cookbook inconsistency identified

The cookbook describes `age_driving_licence` as *"Year in which the insured obtained the driving licence"*, but the observed range is **0–80** with a sample like `{26, 19, 19, 18, 19}`. The variable is in fact the **number of years the licence has been held**. We treat it as years-held throughout this study and rename it `years_licence_held` in the sandbox to prevent downstream confusion.
:::

## Software stack

```{r}
#| label: load-packages

pacman::p_load(
  # Core
  tidyverse, lubridate, scales, knitr, kableExtra,
  # ggplot2 extensions
  ggridges, ggdist, ggrepel, ggstatsplot, ggcorrplot, ggpubr,
  patchwork, ggh4x, ggtext,
  # Interactive
  plotly, ggiraph, crosstalk, DT,
  # Statistics
  performance, FSA
)
```

# Data Wrangling

## Import

The raw file uses **semicolons** as field delimiter and **periods** as decimal mark (European convention). We use `readr::read_delim()` to handle this and explicitly type the columns to avoid `readr`'s heuristics misclassifying e.g. `insured_id` as numeric.

```{r}
#| label: data-import
#| cache: true

raw_path <- "data/insurance_raw.csv"

# Auto-detect delimiter: original Mendeley file is `;`-separated, but the
# project copy may already be comma-normalised. We probe and pick.
first_line <- readr::read_lines(raw_path, n_max = 1)
delim <- if (stringr::str_detect(first_line, ";")) ";" else ","

raw_df <- readr::read_delim(
  raw_path,
  delim = delim,
  locale = readr::locale(decimal_mark = ".", grouping_mark = ""),
  show_col_types = FALSE
)

dim(raw_df)
```

## Audit

A compact data-quality audit before any transformation.

```{r}
#| label: audit
audit_tbl <- raw_df |>
  summarise(across(
    everything(),
    list(
      n_missing = ~sum(is.na(.x)),
      pct_missing = ~mean(is.na(.x)) * 100,
      n_unique = ~dplyr::n_distinct(.x)
    ),
    .names = "{.col}__{.fn}"
  )) |>
  tidyr::pivot_longer(everything(),
                      names_to = c("variable", "metric"),
                      names_sep = "__") |>
  tidyr::pivot_wider(names_from = metric, values_from = value) |>
  arrange(desc(pct_missing))

audit_tbl |>
  filter(pct_missing > 0 | n_unique <= 10) |>
  knitr::kable(
    digits = 3,
    caption = "Variables with missingness or low cardinality",
    col.names = c("Variable", "Missing (n)", "Missing (%)", "Unique values")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

**Findings**:

-   Missingness is negligible: `fuel_type` (0.36%), `vehicle_value` (0.14%), `vehicle_age` (\<0.001%).
-   49 records have `total_premium = 0` **and** `total_exposure = 0`, these are inactive shells and must be removed.
-   `policy_status` has only two levels (A/C); `business_type` two (NB/P); `bonus_score` three (G/N/B).

## Cleaning & derivation

```{r}
#| label: wrangle
#| cache: true

sandbox <- raw_df |>
  # 1) rename the misleading cookbook variable
  rename(years_licence_held = age_driving_licence) |>
  # 2) drop inactive shells (0 premium & 0 exposure)
  filter(!(total_premium == 0 & total_exposure == 0)) |>
  # 3) impute the trivial missingness with category "Unknown"
  mutate(
    fuel_type = forcats::fct_na_value_to_level(factor(fuel_type), level = "Unknown")
  ) |>
  # 4) categorical factors with meaningful levels & labels
  mutate(
    policy_type      = factor(policy_type,
                              levels = c("TP","TPG","CC","COMP_E","COMP_N"),
                              labels = c("TP (basic)","TPG (+glass)",
                                         "CC (combined)","COMP-E (excess)",
                                         "COMP-N (no excess)")),
    policy_status    = factor(policy_status, levels = c("A","C"),
                              labels = c("Active","Cancelled")),
    business_type    = factor(business_type, levels = c("NB","P"),
                              labels = c("New Business","Portfolio")),
    payment_frequency= factor(payment_frequency, levels = c("A","S","Q"),
                              labels = c("Annual","Semi-annual","Quarterly")),
    bonus_score      = factor(bonus_score, levels = c("G","N","B"),
                              labels = c("Good","Neutral","Bad")),
    municipality_type= factor(municipality_type, levels = c("I","C","IS"),
                              labels = c("Inland","Coastal","Islands")),
    circulation_area = factor(circulation_area, levels = c("U","R"),
                              labels = c("Urban","Rural")),
    year_f           = factor(year)
  ) |>
  # 5) derived KPIs - the core analytical variables
  mutate(
    # exposure-adjusted measures
    pure_premium    = total_incurred / pmax(total_exposure, 1e-6),
    claim_frequency = total_claims   / pmax(total_exposure, 1e-6),
    # severity defined only when at least one claim
    severity        = ifelse(total_claims > 0, total_incurred / total_claims, NA_real_),
    # the headline profitability metric
    loss_ratio      = total_incurred / pmax(total_premium, 1e-6),
    # has the policy claimed at all?
    has_claim       = total_claims > 0,
    # large-loss flag (P99 of non-zero incurred ≈ EUR 3,800)
    large_loss      = total_incurred > 3784
  ) |>
  # 6) banded continuous variables for stratified analysis
  mutate(
    driver_age_band = cut(driver_age,
                          breaks = c(17,25,35,45,55,65,Inf),
                          labels = c("18-25","26-35","36-45",
                                     "46-55","56-65","65+"),
                          right  = TRUE),
    vehicle_age_band = cut(vehicle_age,
                           breaks = c(-1,5,10,15,20,30,Inf),
                           labels = c("0-5","6-10","11-15",
                                      "16-20","21-30","30+"),
                           right  = TRUE),
    veh_value_band   = cut(vehicle_value,
                           breaks = c(0,10000,20000,30000,50000,Inf),
                           labels = c("<10k","10-20k","20-30k","30-50k","50k+"),
                           right  = TRUE),
    licence_band     = cut(years_licence_held,
                           breaks = c(-1,5,15,25,35,Inf),
                           labels = c("0-5","6-15","16-25","26-35","35+"),
                           right  = TRUE)
  )

# verify final shape
glue::glue("Sandbox: {nrow(sandbox)} rows × {ncol(sandbox)} cols (added {ncol(sandbox) - ncol(raw_df)} derived vars)")
```

## Glimpse of the analytical sandbox

```{r}
#| label: sandbox-glimpse
sandbox |>
  select(year, policy_type, bonus_score, driver_age, vehicle_age,
         total_premium, total_incurred, total_exposure,
         loss_ratio, pure_premium, claim_frequency) |>
  head(8) |>
  knitr::kable(digits = 2, caption = "First 8 rows of the analytical sandbox") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                font_size = 12)
```

```{r save-sandbox, include=FALSE}
saveRDS(sandbox, "data/analytical_sandbox.rds")
```

# Univariate Visual Analytics

> **Goal of this section** characterise the marginal distributions of every variable that may drive profitability or volatility, before any bivariate analysis. We deliberately separate **numerical** from **categorical** because the right grammar (histogram + density vs. bar + Pareto) differs.

## Numerical variables

### A pair of complementary views: histogram + boxplot

For each principal numeric variable we pair a **binned histogram with overlaid density** (shape) with a **boxplot** (centre/spread/outliers). The combination tells us, in one glance, both the *shape of the bulk* and the *behaviour of the tails*.

```{r}
#| label: fig-num-univariate
#| fig-cap: "Distribution of key numeric drivers. Top: histogram + density. Bottom: boxplot with outliers in red."
#| fig-height: 9

theme_wrc <- function(base_size = 11) {
  theme_minimal(base_size = base_size, base_family = "sans") +
    theme(
      plot.title       = element_text(face = "bold", size = rel(1.05), color = "#111111"),
      plot.subtitle    = element_text(color = "#555555", size = rel(0.9)),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#EAEAEA"),
      axis.text        = element_text(color = "#333333"),
      strip.text       = element_text(face = "bold", color = "#111111"),
      strip.background = element_rect(fill = "#F0F0F0", color = NA)
    )
}

num_vars <- c("driver_age","vehicle_age","years_licence_held",
              "vehicle_value","power_to_weight_ratio","total_premium")

long_num <- sandbox |>
  select(all_of(num_vars)) |>
  tidyr::pivot_longer(everything(), names_to = "var", values_to = "value") |>
  mutate(var = factor(var, levels = num_vars))

p_hist <- ggplot(long_num, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 40, fill = "#222222", color = "white", alpha = 0.85) +
  geom_density(color = "#C8102E", linewidth = 0.7) +
  facet_wrap(~ var, scales = "free", ncol = 3) +
  labs(title = "Marginal distributions of numeric drivers",
       subtitle = "Histogram (density-scaled) with kernel-density overlay",
       x = NULL, y = "Density") +
  theme_wrc()

p_box <- ggplot(long_num, aes(x = value, y = "")) +
  geom_boxplot(outlier.color = "#C8102E", outlier.alpha = 0.4,
               outlier.size = 0.7, fill = "#F5F5F5",
               color = "#222222", width = 0.4) +
  facet_wrap(~ var, scales = "free_x", ncol = 3) +
  labs(x = NULL, y = NULL,
       subtitle = "Boxplot — outliers (≥1.5×IQR) highlighted") +
  theme_wrc() + theme(axis.text.y = element_blank())

(p_hist / p_box) + patchwork::plot_layout(heights = c(2, 1))
```

::: callout-tip
## What we learn

-   **`driver_age`** is roughly Gaussian (mean ≈ 49) : no extreme young-driver mass that one might expect in a UK book.
-   **`vehicle_age`** is **bimodal** with a long right tail past 50 years : Spain has a vintage-vehicle subculture worth investigating.
-   **`total_premium`** is **right-skewed** (median ≈ €258, max \> €5,000). LR analyses on the *raw* scale will be dominated by tiny-premium policies; we will use log scale or robust statistics where appropriate.
-   **`vehicle_value`** has a similar log-normal shape : the high-end tail will overlap with the `power_to_weight_ratio` low tail (sports cars).
:::

### Premium vs. incurred: the two financial spines

The two variables that *define* profitability deserve their own diagnostic. Both are right-skewed and the loss distribution is zero-inflated (88% of policies have no loss). A **log-scale** view of the non-zero values exposes the otherwise-invisible structure.

```{r}
#| label: fig-premium-incurred
#| fig-cap: "Premium and incurred-loss distributions. Left: linear scale. Right: log-scale, non-zero only — revealing log-normal tails."
#| fig-height: 5

p1 <- sandbox |>
  ggplot(aes(x = total_premium)) +
  geom_histogram(bins = 50, fill = "#222222", color = "white", alpha = 0.85) +
  scale_x_continuous(labels = scales::label_dollar(prefix = "€")) +
  labs(title = "Total premium · linear", x = NULL, y = "Policies") +
  theme_wrc()

p2 <- sandbox |>
  filter(total_incurred > 0) |>
  ggplot(aes(x = total_incurred)) +
  geom_histogram(bins = 50, fill = "#C8102E", color = "white", alpha = 0.85) +
  scale_x_log10(labels = scales::label_dollar(prefix = "€")) +
  labs(title = "Total incurred · log scale (non-zero)",
       x = NULL, y = "Policies") +
  theme_wrc()

p1 + p2
```

## Categorical variables

### A Pareto-style sweep

For categorical variables, the question is not "what is the shape?" but "**where does the mass concentrate?**". A Pareto-flavoured horizontal bar chart, ordered by frequency and annotated with the cumulative share, makes concentration immediately legible.

```{r}
#| label: fig-cat-univariate
#| fig-cap: "Frequency of each categorical level. Bars are ordered descending and annotated with within-variable share."
#| fig-height: 8

cat_vars <- c("policy_type","policy_status","business_type",
              "payment_frequency","bonus_score","fuel_type",
              "municipality_type","circulation_area")

cat_summary <- sandbox |>
  select(all_of(cat_vars)) |>
  tidyr::pivot_longer(everything(), names_to = "var", values_to = "level") |>
  count(var, level) |>
  group_by(var) |>
  mutate(share = n / sum(n),
         level = forcats::fct_reorder(as.character(level), n)) |>
  ungroup() |>
  mutate(var = factor(var, levels = cat_vars))

ggplot(cat_summary, aes(x = n, y = level)) +
  geom_col(fill = "#222222", alpha = 0.9, width = 0.7) +
  geom_text(aes(label = scales::percent(share, accuracy = 0.1)),
            hjust = -0.1, size = 3, color = "#333333") +
  facet_wrap(~ var, scales = "free", ncol = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25)),
                     labels = scales::label_number(scale = 1e-3,
                                                   suffix = "k")) +
  labs(title = "Categorical variable composition",
       subtitle = "Bars: count of policy-years · Labels: share within variable",
       x = "Policy-year count (thousands)", y = NULL) +
  theme_wrc() +
  theme(axis.text.y = element_text(size = 9))
```

::: callout-tip
## What we learn

-   **Severe class imbalance** in three variables:
    -   `bonus_score`: 96.5% **Good** : the vast majority enjoy a favourable claims history. This means analyses on Neutral/Bad must be done cautiously due to small subgroup size.
    -   `policy_status`: 89.2% Active.
    -   `business_type`: 69.3% New Business.
-   `policy_type` is dominated by **CC (combined)** at 57.8% with **COMP-N** (the no-excess comprehensive) representing just 4.6%, yet, as we will show below, COMP-N is the single largest profitability risk in the book.
-   Geographic split is roughly even between **Urban (53.8%)** and **Rural (46.2%)**.
:::

### The top-10 vehicle-brand long tail

`vehicle_brand` has 68 unique levels. A faceted bar chart would be illegible, instead, a **lollipop chart of the top 10** is more honest about the long tail.

```{r}
#| label: fig-brand
#| fig-cap: "Top-10 vehicle brands by policy-year count. The remaining 58 brands account for ~28% of the book."
#| fig-height: 4.5

brand_top <- sandbox |>
  count(vehicle_brand, sort = TRUE) |>
  mutate(rank = row_number(),
         share = n / sum(n))

top10 <- brand_top |> slice_head(n = 10)
others <- brand_top |> slice_tail(n = -10)

top10 |>
  mutate(vehicle_brand = forcats::fct_reorder(vehicle_brand, n)) |>
  ggplot(aes(x = n, y = vehicle_brand)) +
  geom_segment(aes(x = 0, xend = n,
                   y = vehicle_brand, yend = vehicle_brand),
               color = "#888888", linewidth = 0.6) +
  geom_point(size = 4, color = "#222222") +
  geom_text(aes(label = scales::percent(share, accuracy = 0.1)),
            hjust = -0.4, size = 3.2, color = "#333333") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = scales::label_number(scale = 1e-3,
                                                   suffix = "k")) +
  annotate("text", x = max(top10$n) * 0.85, y = 1.3,
           label = glue::glue("Long tail: {nrow(others)} more brands · {scales::percent(sum(others$share),0.1)}"),
           hjust = 1, color = "#C8102E", fontface = "italic", size = 3.5) +
  labs(title = "Top-10 vehicle brands dominate but do not exhaust the book",
       subtitle = "Lollipop chart · Labels show share of all policy-years",
       x = "Policy-year count (thousands)", y = NULL) +
  theme_wrc()
```

# Bivariate Visual Analytics — Profitability Drivers

> **For** each candidate driver, ask **"does it move the loss ratio?"** and validate the visual difference with a statistical test. We use **portfolio-weighted LR** (`Σ incurred / Σ premium`) for headline numbers and **policy-level LR** for statistical testing.

## Portfolio-LR comparison across categorical drivers

A single small-multiple chart compares the headline profitability metric across every categorical driver simultaneously. The horizontal reference line marks the **overall portfolio LR (≈ 70%)**, which acts as a "break-even" sense check.

```{r}
#| label: fig-lr-by-cat
#| fig-cap: "Portfolio Loss Ratio by category. Bars above the dashed line are worse than the portfolio average; below are better."
#| fig-height: 10

portfolio_lr <- with(sandbox, sum(total_incurred) / sum(total_premium))

lr_by_cat <- function(df, var) {
  df |>
    group_by(level = .data[[var]]) |>
    summarise(
      premium  = sum(total_premium),
      incurred = sum(total_incurred),
      n        = dplyr::n(),
      .groups  = "drop"
    ) |>
    mutate(LR = incurred / premium, var = var,
           level = as.character(level))
}

cat_for_lr <- c("policy_type","business_type","payment_frequency",
                "bonus_score","fuel_type","municipality_type",
                "circulation_area","policy_status")

lr_panel <- purrr::map_dfr(cat_for_lr, ~lr_by_cat(sandbox, .x)) |>
  group_by(var) |>
  mutate(level = forcats::fct_reorder(level, LR)) |>
  ungroup() |>
  mutate(diff_pp = (LR - portfolio_lr) * 100,
         bar_col = ifelse(LR > portfolio_lr, "#C8102E", "#2E7D32"))

ggplot(lr_panel, aes(x = LR, y = level, fill = bar_col)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_vline(xintercept = portfolio_lr,
             linetype = "dashed", color = "#222222", linewidth = 0.4) +
  geom_text(aes(label = scales::percent(LR, accuracy = 0.1)),
            hjust = -0.1, size = 3, color = "#333333") +
  scale_fill_identity() +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ var, scales = "free_y", ncol = 2) +
  labs(title = "Loss Ratio by categorical driver",
       subtitle = glue::glue("Dashed line: overall portfolio LR ({scales::percent(portfolio_lr, 0.1)}) · Red = worse, Green = better"),
       x = "Loss Ratio (Σ incurred / Σ premium)", y = NULL) +
  theme_wrc() +
  theme(strip.text = element_text(face = "bold"))
```

::: callout-important
## Three findings stand out

1.  **`policy_type`** has by far the widest LR spread from 54% (TP basic) to 86% (COMP-N no-excess). The product mix is the *first lever* for portfolio remediation.
2.  **`bonus_score`** is *not* monotonic: **Good (70%) \< Bad (79%) \< Neutral (81%)**. The Neutral pool appears to be the dumping-ground for borderline risks the bonus system cannot classify cleanly, a candidate underwriting investigation.
3.  **Diesel vs. Gasoline**: Diesel runs **9 pp higher** LR than Gasoline. With \~226k diesel records the gap is highly significant (validated below) and reflects either selection (diesel drivers cover more km) or rating-table miscalibration.
:::

## Statistical validation with `ggstatsplot`

A bar of a 9-point difference is meaningless without knowing whether it could arise by chance. We use `ggstatsplot::ggbetweenstats()`, which embeds the test result (Kruskal-Wallis for non-normal continuous outcomes) **directly into the plot title**, exactly the kind of audit trail a visual-analytics deliverable needs.

```{r}
#| label: fig-stat-policy-type
#| fig-cap: "Policy-level LR distribution by policy_type with Kruskal-Wallis test. The non-parametric test is appropriate because LR is heavily zero-inflated and right-skewed."
#| fig-height: 6

# subsample for stat plot — 50k is plenty for a stable test result
set.seed(608)
stat_sample <- sandbox |>
  filter(total_premium > 0, loss_ratio < 5) |>  # cap outliers at 500% LR for visibility
  slice_sample(n = 50000)

ggstatsplot::ggbetweenstats(
  data = stat_sample,
  x = policy_type,
  y = loss_ratio,
  type = "nonparametric",
  pairwise.display = "significant",
  centrality.plotting = TRUE,
  centrality.type = "nonparametric",
  bf.message = FALSE,
  ggtheme = theme_wrc(),
  title = "Loss-ratio dispersion by policy type — non-parametric test",
  xlab = NULL,
  ylab = "Policy-level Loss Ratio"
) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 3))
```

## Continuous driver: driver_age × LR with a U-shape

Driver age is the canonical actuarial risk variable. We expect a **U-shape**: young drivers (inexperience) and elderly drivers (slower reflexes) drive higher LR. We test this by combining a **binned column** (band-level portfolio LR) with a **smoothed line** (policy-level LOESS).

```{r}
#| label: fig-driver-age-u
#| fig-cap: "Loss ratio by driver age: bands (bars, portfolio-weighted) vs. smoothed line (LOESS on policy-level). The smoothed line reveals a flat profile, contradicting the classical U-shape — pricing has already absorbed the age signal."
#| fig-height: 5.5

age_lr_band <- sandbox |>
  group_by(driver_age_band) |>
  summarise(LR = sum(total_incurred) / sum(total_premium),
            n = dplyr::n(), .groups = "drop") |>
  filter(!is.na(driver_age_band))

age_lr_cont <- sandbox |>
  filter(driver_age >= 18, driver_age <= 90,
         total_premium > 0, loss_ratio < 3)

ggplot() +
  geom_col(data = age_lr_band,
           aes(x = driver_age_band, y = LR),
           fill = "#222222", alpha = 0.85, width = 0.7) +
  geom_text(data = age_lr_band,
            aes(x = driver_age_band, y = LR,
                label = scales::percent(LR, 0.1)),
            vjust = -0.5, size = 3.2, color = "#333333") +
  geom_hline(yintercept = portfolio_lr,
             linetype = "dashed", color = "#C8102E") +
  annotate("text", x = 5.5, y = portfolio_lr + 0.02,
           label = glue::glue("Portfolio LR = {scales::percent(portfolio_lr, 0.1)}"),
           color = "#C8102E", hjust = 1, size = 3.2, fontface = "italic") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, max(age_lr_band$LR) * 1.2)) +
  labs(title = "Driver age vs. Loss Ratio — the U-shape is absent",
       subtitle = "Portfolio-weighted LR by age band · Dashed line: book average",
       x = "Driver age band", y = "Loss Ratio") +
  theme_wrc()
```

::: callout-note
## Counter-intuitive finding

Conventional actuarial wisdom predicts a **U-shape** with elevated LR for the youngest and oldest drivers. Here the profile is essentially **flat at 67–73%** likely because the insurer's premium-rating engine has already loaded young/old drivers correctly, neutralising the gross-of-pricing effect. In other words, **the absence of a U-shape is evidence of effective age-based pricing**, not of an absence of age-related risk.
:::

## Vehicle age & vehicle value — the diagonal

```{r}
#| label: fig-vehicle-heatmap
#| fig-cap: "Heatmap of Loss Ratio by vehicle age band × vehicle value band. The darkest cells in the upper-left (new + expensive) flag the highest-risk segment."
#| fig-height: 5

heat <- sandbox |>
  filter(!is.na(vehicle_age_band), !is.na(veh_value_band)) |>
  group_by(vehicle_age_band, veh_value_band) |>
  summarise(LR = sum(total_incurred) / sum(total_premium),
            n  = dplyr::n(), .groups = "drop") |>
  filter(n >= 100)  # suppress tiny cells

ggplot(heat, aes(x = veh_value_band, y = vehicle_age_band, fill = LR)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(n > 100,
                               scales::percent(LR, accuracy = 0.1), "")),
            color = "white", fontface = "bold", size = 3.3) +
  scale_fill_gradient(low = "#9CCC65", high = "#B71C1C",
                      labels = scales::percent_format(accuracy = 1),
                      name = "Loss Ratio") +
  labs(title = "Vehicle profile heatmap — where the risk concentrates",
       subtitle = "Cells with fewer than 100 policy-years suppressed",
       x = "Vehicle value band", y = "Vehicle age band") +
  theme_wrc() +
  theme(panel.grid = element_blank())
```

## Frequency-Severity decomposition

A classic actuarial decomposition: **Loss Cost = Frequency × Severity**. Visualising the two on independent axes (a "Frequency-Severity quadrant chart") locates each policy_type in a 2-D risk space.

```{r}
#| label: fig-freq-sev
#| fig-cap: "Frequency-severity quadrant by policy type. Bubble size = portfolio share; colour = loss ratio."
#| fig-height: 5

fs <- sandbox |>
  filter(total_exposure > 0) |>
  group_by(policy_type) |>
  summarise(
    freq      = sum(total_claims) / sum(total_exposure),
    severity  = sum(total_incurred[total_claims > 0], na.rm = TRUE) /
                pmax(sum(total_claims[total_claims > 0], na.rm = TRUE), 1),
    LR        = sum(total_incurred) / sum(total_premium),
    n         = dplyr::n(),
    .groups   = "drop"
  )

med_f <- median(fs$freq); med_s <- median(fs$severity)

ggplot(fs, aes(x = freq, y = severity)) +
  geom_hline(yintercept = med_s, linetype = "dashed", color = "#888888") +
  geom_vline(xintercept = med_f, linetype = "dashed", color = "#888888") +
  geom_point(aes(size = n, fill = LR), shape = 21, color = "#222222",
             stroke = 0.8) +
  ggrepel::geom_text_repel(aes(label = policy_type),
                           size = 3.5, fontface = "bold",
                           color = "#111111", box.padding = 0.6,
                           segment.color = "#999999") +
  scale_size_continuous(range = c(6, 22), guide = "none") +
  scale_fill_gradient(low = "#9CCC65", high = "#B71C1C",
                      labels = scales::percent_format(accuracy = 1),
                      name = "LR") +
  scale_x_continuous(labels = scales::number_format(accuracy = 0.01)) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(title = "Frequency × Severity decomposition by policy type",
       subtitle = "Bubble area = policy-year count · Dashed lines: median frequency & severity",
       x = "Claim frequency (claims per year of exposure)",
       y = "Average severity (€ per claim)") +
  theme_wrc()
```

::: callout-tip
## Reading the quadrant

-   **COMP-N** is high on **both** axes: it claims more often *and* each claim is bigger. It is the only type sitting in the "double trouble" upper-right quadrant.
-   **TP (basic)** is on the opposite corner: rare claims that cost little. The lowest LR in the book.
-   **CC (combined)** is the volume leader (largest bubble) and sits near the median on both axes, it sets the portfolio's centre of gravity.
:::

## Correlation map of numeric drivers

A single correlation heatmap of all numeric drivers (vs. each other and vs. profitability KPIs) acts as a navigation overview.

```{r}
#| label: fig-corr
#| fig-cap: "Spearman correlation matrix among numeric drivers and KPIs. Spearman (rank) used because most distributions are heavily right-skewed."
#| fig-height: 6

num_for_corr <- sandbox |>
  select(driver_age, vehicle_age, years_licence_held, vehicle_value,
         power_to_weight_ratio, seats, total_premium,
         total_claims, total_incurred, total_exposure, loss_ratio) |>
  filter(if_all(everything(), ~ !is.na(.)))

cor_mat <- cor(num_for_corr, method = "spearman")

ggcorrplot::ggcorrplot(
  cor_mat,
  hc.order = TRUE, type = "lower",
  lab = TRUE, lab_size = 3,
  colors = c("#B71C1C", "#FFFFFF", "#2E7D32"),
  outline.color = "white"
) +
  labs(title = "Spearman correlations among drivers and KPIs",
       subtitle = "Hierarchical clustering groups co-varying drivers") +
  theme_wrc() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

# Bivariate Visual Analytics — Volatility & Tail Risk

> **Q**uantify *how unstable* each segment is, in addition to *how unprofitable*. Two segments can share a 70% LR but one may swing between 30%-180% policy-by-policy while the other clusters tightly. The first is uninsurable without reinsurance; the second is a stable book.

## Coefficient-of-variation comparison

The **CV (= sd/mean)** of policy-level LR within each segment is our primary volatility measure. We complement it with the mean LR on the same chart so the analyst can see the **risk-return trade-off** at a glance.

```{r}
#| label: fig-cv
#| fig-cap: "Mean LR vs. CV of LR by policy type. Bottom-left = ideal (low LR, stable). Top-right = avoid (high LR, unstable)."
#| fig-height: 4.5

vol <- sandbox |>
  filter(total_premium > 0, loss_ratio < 10) |>  # 1000% LR cap
  group_by(policy_type) |>
  summarise(
    mean_lr = mean(loss_ratio),
    sd_lr   = sd(loss_ratio),
    cv      = sd_lr / mean_lr,
    n       = dplyr::n(),
    .groups = "drop"
  )

ggplot(vol, aes(x = mean_lr, y = cv, size = n)) +
  geom_point(shape = 21, fill = "#222222", color = "white", stroke = 1) +
  ggrepel::geom_text_repel(aes(label = policy_type),
                           size = 3.5, fontface = "bold",
                           color = "#111111", box.padding = 0.6) +
  scale_size_continuous(range = c(5, 20), guide = "none") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Profitability vs. Volatility frontier",
       subtitle = "X = mean policy LR · Y = coefficient of variation · Bubble = exposure",
       x = "Mean policy-level LR",
       y = "Coefficient of variation (sd / mean)") +
  theme_wrc()
```

## Lorenz curve & Gini — the tail-concentration signature

The most evocative way to show tail concentration is the **Lorenz curve**: sort policies by loss size, plot cumulative population share against cumulative loss share. A diagonal line means perfect equality; a sharp bow towards the bottom-right means a small minority drives most of the loss.

```{r}
#| label: fig-lorenz
#| fig-cap: "Lorenz curve of incurred losses across the entire portfolio. The Gini coefficient summarises the curvature."
#| fig-height: 5.5

lorenz_df <- sandbox |>
  arrange(total_incurred) |>
  mutate(
    cum_pop = row_number() / dplyr::n(),
    cum_loss = cumsum(total_incurred) / sum(total_incurred)
  )

# Gini using trapezoidal area
gini <- 1 - 2 * sum((lorenz_df$cum_pop - dplyr::lag(lorenz_df$cum_pop, default = 0)) *
                    (lorenz_df$cum_loss + dplyr::lag(lorenz_df$cum_loss, default = 0)) / 2)

# Down-sample for plotting (full 354k is overkill for a curve)
lorenz_plot <- lorenz_df |>
  slice(round(seq(1, dplyr::n(), length.out = 5000)))

ggplot(lorenz_plot, aes(x = cum_pop, y = cum_loss)) +
  geom_ribbon(aes(ymin = cum_pop, ymax = cum_loss),
              fill = "#C8102E", alpha = 0.18) +
  geom_abline(slope = 1, intercept = 0,
              color = "#888888", linetype = "dashed") +
  geom_line(color = "#222222", linewidth = 1.1) +
  annotate("text", x = 0.55, y = 0.18,
           label = glue::glue("Gini coefficient = {round(gini, 3)}"),
           color = "#C8102E", fontface = "bold", size = 4.5) +
  annotate("segment", x = 0.95, xend = 0.99,
           y = 0.50, yend = 0.83,
           arrow = arrow(length = unit(0.2,"cm")),
           color = "#222222") +
  annotate("text", x = 0.78, y = 0.45,
           label = "Top 1% of policies\n→ ~50% of losses",
           color = "#222222", hjust = 0, size = 3.6, fontface = "italic") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Lorenz curve — extreme loss concentration",
       subtitle = "Shaded area = inequality between population and loss shares",
       x = "Cumulative share of policies (ascending by loss)",
       y = "Cumulative share of losses") +
  theme_wrc()
```

::: callout-important
## A near-catastrophic tail

A Gini around **0.95** is extraordinary even by P&C standards (a typical auto-book sits at 0.85–0.92). It tells us that **the volatility of this portfolio is essentially the volatility of a few-hundred policies**. Any portfolio-level management action must specifically address the upper tail (e.g., excess-of-loss reinsurance, per-risk underwriting limits) rather than blanket re-rating.
:::

## Severity distribution by segment — raincloud plot

A **raincloud plot** (`ggdist`) layers a half-violin, a boxplot and the raw points, three encodings of the same data, each catching what the others miss.

```{r}
#| label: fig-severity-raincloud
#| fig-cap: "Severity distribution by policy type (log scale). Raincloud = half-violin (shape) + boxplot (centre/spread) + jittered points (raw data)."
#| fig-height: 6

sev_df <- sandbox |>
  filter(total_claims > 0, severity > 0, severity < 1e5) |>
  select(policy_type, severity)

ggplot(sev_df, aes(x = policy_type, y = severity, fill = policy_type)) +
  ggdist::stat_halfeye(adjust = 0.6, width = 0.55,
                       .width = 0, justification = -0.22, alpha = 0.85) +
  geom_boxplot(width = 0.18, outlier.shape = NA,
               alpha = 0.7, color = "#111111") +
  ggdist::stat_dots(side = "left", justification = 1.12,
                    binwidth = NA, dotsize = 0.4,
                    color = "#666666", alpha = 0.25) +
  scale_y_log10(labels = scales::dollar_format(prefix = "€")) +
  scale_fill_manual(values = c("#90A4AE","#78909C","#546E7A",
                               "#37474F","#263238"), guide = "none") +
  labs(title = "Claim severity by policy type",
       subtitle = "Log-scale · Half-violin (top) · Boxplot (centre) · Dot-density (bottom)",
       x = NULL, y = "Severity per claim (log scale)") +
  theme_wrc() +
  coord_flip()
```

## Density ridges — distributional shifts by bonus class

```{r}
#| label: fig-ridges
#| fig-cap: "Loss-ratio density by bonus_score (zoomed to LR ≤ 200%). The Bad class shows a thicker right tail; the Good class is sharply zero-inflated."
#| fig-height: 4.5

ridge_df <- sandbox |>
  filter(total_premium > 0, loss_ratio < 2) |>
  select(bonus_score, loss_ratio)

ggplot(ridge_df, aes(x = loss_ratio, y = bonus_score, fill = bonus_score)) +
  ggridges::geom_density_ridges(scale = 1.4, alpha = 0.85,
                                color = "white", linewidth = 0.4,
                                rel_min_height = 0.01) +
  scale_fill_manual(values = c("Good" = "#2E7D32",
                               "Neutral" = "#888888",
                               "Bad" = "#C8102E"),
                    guide = "none") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Distribution of policy-level LR by bonus class",
       subtitle = "Trimmed at LR ≤ 200% for readability",
       x = "Policy-level Loss Ratio", y = NULL) +
  theme_wrc()
```

# Temporal Visual Analytics

## Three-year trend — diagnosing the deterioration

LR is climbing year-on-year. A natural-decomposition view rebases the three multiplicative components, frequency, severity, and *premium adequacy* to 100 in 2022. Whichever line moves the most is the culprit.

```{r}
#| label: fig-trend
#| fig-cap: "Where the deterioration comes from: indexed evolution of LR's components. Frequency and severity are essentially flat; premium-per-exposure is in free-fall."
#| fig-height: 5

trend <- sandbox |>
  group_by(year) |>
  summarise(
    LR        = sum(total_incurred) / sum(total_premium),
    frequency = sum(total_claims) / sum(total_exposure),
    severity  = sum(total_incurred[total_claims > 0]) /
                pmax(sum(total_claims[total_claims > 0]), 1),
    prem_per_exp = sum(total_premium) / sum(total_exposure),
    pure_prem    = sum(total_incurred) / sum(total_exposure),
    .groups   = "drop"
  ) |>
  tidyr::pivot_longer(-year, names_to = "metric", values_to = "value") |>
  group_by(metric) |>
  mutate(index = value / value[year == 2022] * 100) |>
  ungroup() |>
  mutate(metric = factor(metric,
    levels = c("LR","prem_per_exp","pure_prem","frequency","severity"),
    labels = c("Loss Ratio","Premium / exposure",
               "Pure premium / exposure","Claim frequency","Severity")))

ggplot(trend, aes(x = year, y = index, color = metric, group = metric)) +
  geom_hline(yintercept = 100, color = "#888888", linetype = "dashed") +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3.5) +
  geom_text(data = trend |> filter(year == 2024),
            aes(label = paste0(metric, " (", scales::number(index, 1), ")")),
            hjust = -0.05, fontface = "bold", size = 3.3) +
  scale_color_manual(values = c("Loss Ratio" = "#C8102E",
                                "Premium / exposure" = "#1565C0",
                                "Pure premium / exposure" = "#FF8F00",
                                "Claim frequency" = "#888888",
                                "Severity" = "#222222")) +
  scale_x_continuous(breaks = 2022:2024,
                     limits = c(2022, 2025.3),
                     expand = expansion(mult = c(0.02, 0))) +
  labs(title = "Diagnosing the LR climb — it is a pricing problem",
       subtitle = "All metrics rebased to 100 in 2022 · Premium per exposure has fallen 9%",
       x = NULL, y = "Index (2022 = 100)",
       color = NULL) +
  theme_wrc() +
  theme(legend.position = "none")
```

::: callout-important
## The deterioration is a pricing story, not a loss-cost story

-   **Premium per exposure** fell from €446 to €405, a **9% decline** over two years. The book is being written for less money per unit of risk.
-   **Pure premium per exposure** dipped in 2023 (-3%) then rebounded to \~€302 (+3% vs 2022), net stable.
-   **Frequency** and **severity** moved by 3% and 0% respectively over the window.
-   Conclusion: the 14% jump in LR is overwhelmingly attributable to **inadequate premium growth** (most likely competitive market pressure), not to claims inflation. This re-frames the remediation playbook from "tighten claims handling" to "**restore pricing discipline**".
:::

## Year × policy-type heatmap — where the rot starts

```{r}
#| label: fig-year-policy-heat
#| fig-cap: "LR by year × policy type. COMP-N crosses the 100% break-even line in 2024 — an underwriting alarm."
#| fig-height: 4.5

yp <- sandbox |>
  group_by(year, policy_type) |>
  summarise(LR = sum(total_incurred) / sum(total_premium),
            .groups = "drop")

ggplot(yp, aes(x = factor(year), y = policy_type, fill = LR)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = scales::percent(LR, 0.1),
                color = ifelse(LR > 0.85, "white", "#222222")),
            fontface = "bold", size = 3.6) +
  scale_color_identity() +
  scale_fill_gradient2(low = "#2E7D32", mid = "#FFF59D", high = "#B71C1C",
                       midpoint = 0.75,
                       labels = scales::percent_format(accuracy = 1),
                       name = "LR") +
  labs(title = "Loss-ratio trajectory · COMP-N enters unprofitable territory in 2024",
       subtitle = "Diverging colour scale centred on 75% (a typical break-even target after expenses)",
       x = NULL, y = NULL) +
  theme_wrc() +
  theme(panel.grid = element_blank())
```

# Bonus — Interactive Visual Analytics

> A static figure communicates an answer the analyst already found. An interactive figure lets the *reader* find a different answer. The three widgets below cover three distinct interaction patterns: **hover-and-inspect**, **linked brushing**, and **drill-down exploration**.

## Bonus 1 · Interactive 3-D heatmap with year drill-down (`plotly`)

The LR by `policy_type × bonus_score` heatmap, faceted by year. Hover any cell to read the exact LR, exposure and incurred loss; click a year tab to switch.

```{r}
#| label: bonus-1-plotly-heatmap
#| fig-cap: "Interactive heatmap — hover any cell for the exact figures; toggle years using the dropdown."

cube <- sandbox |>
  group_by(year, policy_type, bonus_score) |>
  summarise(
    n         = dplyr::n(),
    premium   = sum(total_premium),
    incurred  = sum(total_incurred),
    LR        = incurred / premium,
    .groups   = "drop"
  ) |>
  mutate(LR = pmin(LR, 2))   # cap colour scale at 200% for legibility

p_heat <- plotly::plot_ly(
  data = cube,
  x = ~policy_type,
  y = ~bonus_score,
  z = ~LR,
  frame = ~year,
  type = "heatmap",
  colorscale = list(c(0, "#2E7D32"),
                    c(0.4, "#FFF59D"),
                    c(1, "#B71C1C")),
  zmin = 0, zmax = 2,
  hovertemplate = paste0(
    "<b>Policy type:</b> %{x}<br>",
    "<b>Bonus class:</b> %{y}<br>",
    "<b>LR:</b> %{z:.1%}<br>",
    "<extra></extra>"
  )
) |>
  plotly::layout(
    title = list(text = "Loss Ratio cube — explore by year",
                 font = list(size = 14, color = "#111111")),
    xaxis = list(title = "Policy type", tickangle = -25),
    yaxis = list(title = "Bonus class"),
    margin = list(t = 60, b = 80),
    paper_bgcolor = "#FFFFFF",
    plot_bgcolor = "#FFFFFF"
  ) |>
  plotly::animation_opts(frame = 800, transition = 400, redraw = TRUE) |>
  plotly::animation_slider(currentvalue = list(prefix = "Year: ",
                                               font = list(color = "#111111")))

p_heat
```

## Bonus 2 · Linked brushing — premium vs. incurred (`crosstalk` + `plotly`)

Two views of the **same** policies, a scatter and a per-policy-type LR bar chart. Drag a rectangle on the scatter and the bar chart automatically recalculates LR using only the selected policies. This is the canonical "see-and-discover" interaction of visual analytics.

```{r}
#| label: bonus-2-crosstalk
#| fig-cap: "Linked brushing demo — drag a rectangle on the scatter; the bar chart on the right responds in real time."
#| fig-height: 5.5

# Down-sample for performance — 8k claimant policies is plenty for the visual
set.seed(42)
brush_df <- sandbox |>
  filter(total_claims > 0, total_premium > 0,
         total_incurred < 1e5, total_premium < 2000) |>
  slice_sample(n = 8000) |>
  select(policy_type, bonus_score, year,
         total_premium, total_incurred, loss_ratio)

shared <- crosstalk::SharedData$new(brush_df, key = ~ seq_len(nrow(brush_df)))

scat <- plotly::plot_ly(shared,
                        x = ~total_premium, y = ~total_incurred,
                        color = ~policy_type,
                        colors = c("#90A4AE","#78909C","#546E7A",
                                   "#37474F","#263238"),
                        type = "scatter", mode = "markers",
                        marker = list(size = 5, opacity = 0.55),
                        hovertemplate = paste0(
                          "Premium: €%{x:.0f}<br>",
                          "Incurred: €%{y:.0f}<br>",
                          "Type: %{customdata}<extra></extra>"
                        ),
                        customdata = ~policy_type,
                        width = 460, height = 420) |>
  plotly::layout(
    title = "Premium vs. Incurred — drag to brush",
    xaxis = list(title = "Total premium (€)"),
    yaxis = list(title = "Total incurred (€)"),
    dragmode = "select",
    paper_bgcolor = "#FFFFFF", plot_bgcolor = "#FFFFFF"
  ) |>
  plotly::highlight(on = "plotly_selected", off = "plotly_deselect",
                    persistent = FALSE, dynamic = FALSE,
                    color = "#C8102E", opacityDim = 0.18)

bar <- plotly::plot_ly(shared,
                       x = ~loss_ratio, y = ~policy_type,
                       type = "box", orientation = "h",
                       boxpoints = FALSE,
                       fillcolor = "rgba(34,34,34,0.6)",
                       line = list(color = "#222222"),
                       width = 460, height = 420) |>
  plotly::layout(title = "Selection LR distribution",
                 xaxis = list(title = "LR", range = c(0, 5)),
                 yaxis = list(title = NULL),
                 paper_bgcolor = "#FFFFFF", plot_bgcolor = "#FFFFFF")

crosstalk::bscols(scat, bar)
```

## Bonus 3 · Tooltip-rich drill-down (`ggiraph`)

`ggiraph` makes every ggplot geom interactive without sacrificing the static-plot aesthetics. Here, every brand-point on the lollipop reveals its LR and exposure on hover, and clicking opens a filterable underlying data table.

```{r}
#| label: bonus-3-ggiraph
#| fig-cap: "Hover any brand for its loss ratio. Brand-LR is colour-encoded: red = above portfolio average, green = below."
#| fig-height: 6

brand_lr <- sandbox |>
  group_by(vehicle_brand) |>
  summarise(
    n         = dplyr::n(),
    premium   = sum(total_premium),
    incurred  = sum(total_incurred),
    LR        = incurred / premium,
    .groups   = "drop"
  ) |>
  filter(n >= 1000) |>            # only brands with stable sample
  arrange(desc(n)) |>
  slice_head(n = 20) |>
  mutate(
    brand = forcats::fct_reorder(vehicle_brand, LR),
    above = LR > portfolio_lr,
    tip   = paste0(
      "<b>", vehicle_brand, "</b><br>",
      "Policy-years: ", scales::comma(n), "<br>",
      "Premium: €", scales::comma(round(premium)), "<br>",
      "Incurred: €", scales::comma(round(incurred)), "<br>",
      "LR: ", scales::percent(LR, 0.1)
    )
  )

gg <- ggplot(brand_lr, aes(x = LR, y = brand)) +
  geom_vline(xintercept = portfolio_lr, linetype = "dashed", color = "#888888") +
  geom_segment(aes(x = portfolio_lr, xend = LR,
                   y = brand, yend = brand,
                   color = above), linewidth = 0.6) +
  ggiraph::geom_point_interactive(
    aes(size = n, color = above, tooltip = tip, data_id = brand),
    alpha = 0.85
  ) +
  scale_color_manual(values = c("TRUE" = "#C8102E", "FALSE" = "#2E7D32"),
                     guide = "none") +
  scale_size_continuous(range = c(3, 10), guide = "none") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Vehicle brand × Loss Ratio (top 20 by volume)",
       subtitle = "Hover for details · Bubble = policy-year exposure · Dashed line = portfolio LR",
       x = "Loss Ratio", y = NULL) +
  theme_wrc()

ggiraph::girafe(ggobj = gg, width_svg = 9, height_svg = 6,
                options = list(
                  ggiraph::opts_hover(css = "stroke:#111111;stroke-width:2px;"),
                  ggiraph::opts_tooltip(css = "background:#222222;color:#ffffff;padding:8px;border-radius:4px;font-size:11px;"),
                  ggiraph::opts_toolbar(saveaspng = TRUE)
                ))
```

# Key Findings & Recommendations

::: panel-tabset
## Profitability

1.  **Product mix dominates everything.** `policy_type` drives a 39-pp LR spread (TP at 54% → COMP-N at 94%). Any portfolio remediation should start by re-pricing or de-emphasising COMP-N before anything else, it has already crossed the **break-even line in 2024 (102%)**.
2.  **Cancelled policies bleed money.** Cancelled-status policies show an LR of **128%** versus **64%** for Active ones, and a 1.7× higher claim frequency. The pattern is consistent with **adverse retention**: high-loss policyholders who get cancelled (or self-cancel after claiming) leave behind a disproportionate share of incurred cost. Worth investigating cancellation rules and policy-recapture flows.
3.  **The bonus-malus signal is leaking.** The Neutral class has higher LR (81%) than even the Bad class (79%). This suggests the rule that places policies into Neutral is over-inclusive, it is absorbing risks that should have been classified Bad.
4.  **Diesel runs 9 pp hotter than gasoline.** A statistically rock-solid finding (n ≈ 226k vs. 126k) pointing to either selection (diesel drivers cover more km) or a rating gap.
5.  **Payment-frequency selection.** Annual payers run LR = 68% while semi-annual = 81% and quarterly = 76%. Payment instalment is a proxy for cash-flow stress, which correlates with claims experience.
6.  **Driver age has no residual signal.** The flat LR-by-age profile (67–73% across all bands) is a quiet success of the existing pricing engine.

## Volatility

1.  **The book is a tail-risk machine.** Gini ≈ **0.956**; the top 1% of policies generate \~50% of total losses; the top 0.1% generate \~19%. Reinsurance design (excess-of-loss) is the only viable management tool, re-pricing the body of the book cannot fix what the tail is doing.
2.  **COMP-E carries the highest CV**, i.e. *most volatile* within-segment, even though its mean LR (74%) is only second-worst. It is the segment where pricing precision matters most.
3.  **The deterioration is a *pricing*, not a *loss-cost* story.** Across 2022–2024, premium per exposure fell **9%** while pure premium per exposure was net flat (+3%). The 14-point LR climb is overwhelmingly attributable to softening prices, *not* claims inflation. This re-frames the remediation playbook from "tighten claims handling" to "**restore pricing discipline and re-rate aggressively**".

## Recommended next steps

-   **GLM modelling** of LR against the validated drivers (`policy_type`, `fuel_type`, `bonus_score`, `payment_frequency`, `year`) to produce rebalanced rate factors.
-   **Tail-risk segmentation** : identify the structural characteristics of policies in the top-1% loss bucket and consider per-policy underwriting limits or facultative reinsurance.
-   **Cancellation forensics** : analyse the sequence of claim → cancellation events to isolate insurer-initiated vs. policyholder-initiated terminations and estimate the remediable loss.
-   **Bonus-class redesign** : investigate whether the Neutral class can be split into Neutral-Up and Neutral-Down using rolling claim history.
-   **Pricing health monitor** : adopt a quarterly KPI of *Premium / Exposure* on a rolling-12-month basis; the 2024 erosion would have been flagged six months earlier with this monitor in place.
:::

# References {.unnumbered}

-   *Advanced analytics: unlocking new frontiers in P&C insurance.* (industry report, opening epigraph).
-   Wickham, H. et al. (2023). *Welcome to the tidyverse*. JOSS.
-   Patil, I. (2021). *Visualizations with statistical details: The 'ggstatsplot' approach.* JOSS.
-   Gohel, D. (2024). *ggiraph: Make 'ggplot2' Graphics Interactive.* R package.
-   Sievert, C. (2020). *Interactive Web-Based Data Visualization with R, plotly, and shiny.* Chapman and Hall/CRC.

::: {style="text-align:center;margin-top:3em;color:#888888;font-size:.85em;"}
*ISSS608 Visual Analytics & Applications · Take-home Exercise 1 · `r format(Sys.Date(),'%B %Y')`*
:::
